---
title: "Sample_Import_Preprocessing after cellbender and scrublet"
output: html_document
date: "2024-12-24"
---

Import and process myelin-depleted and control mechanically dissociated single cell 5' gene expression sequenced GBM samples. Input is cellbender output with fpr 0.0. This does not automatically filter any cells. Cells here will be filtered with scrublet outputs (Script 1, python). 

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(readr)
library(stringr)
library(patchwork)
library(rhdf5)
library(SeuratDisk)
library(scCustomize)
library(Matrix)
library(scales)

```

# Load in functions
```{r}
qc_filter_seurat <- function(srt_obj,
                             sample_name = "sample",
                             min_features = 500,
                             max_features = 8000,
                             min_counts = 1000,
                             max_counts = 50000,
                             max_percent_mt = 10,
                             output_dir = "~/results/qc/") {
  
  print(paste0("Running QC for: ", sample_name))
  
  # Calculate percent mitochondrial gene content
  print("calculating percent mt")
  srt_obj[["percent.mt"]] <- PercentageFeatureSet(srt_obj, pattern = "^MT-")
  
  # --- Optional: Remove outliers from nCount_RNA for plotting ---
  print("removing outliers from nCount_RNA for plotting")
  counts <- srt_obj$nCount_RNA
  upper_bound <- quantile(counts, 0.99)
  plot_obj <- subset(srt_obj, subset = nCount_RNA < upper_bound)
  
  # --- Pre-filter Violin Plot ---
  print("making qc violin plot pre-filter")
  vln_list <- lapply(c("nFeature_RNA", "nCount_RNA", "percent.mt"), function(f) {
    VlnPlot(plot_obj, features = f, pt.size = 0.3, alpha = 0.1) +
    scale_y_continuous(breaks = pretty_breaks(n = 7)) +
    theme(legend.position = "none") 
  })
  
  # Combine with patchwork
  print("combining violin plots pre-filter")
  pre_vln <- wrap_plots(vln_list, ncol = 3) +
  plot_annotation(title = paste0("Pre-filter: ", sample_name))

  # --- Apply filtering ---
  print("unfiltered qc data:")
  print(paste0("nFeatures_RNA: ", mean(srt_obj$nFeature_RNA)))
  print(paste0("nCount_RNA: ", mean(srt_obj$nCount_RNA)))
  print(paste0("percent.mt: ", mean(srt_obj$percent.mt)))
  
  print("Filter params:")
  print(paste0("min_features: ", mean(min_features)))
  print(paste0("max_features: ", mean(max_features)))
  print(paste0("min_counts: ", mean(min_counts)))
  print(paste0("max_counts: ", mean(max_counts)))
  print(paste0("max_percent_mt: ", mean(max_percent_mt)))
  
  print("filtering seurat objects")
  srt_filtered <- subset(srt_obj,
                         subset = nFeature_RNA > min_features &
                                  nFeature_RNA < max_features &
                                  nCount_RNA > min_counts &
                                  nCount_RNA < max_counts &
                                  percent.mt < max_percent_mt)
  
  # --- Post-filter Violin Plot ---
  print("making qc violin plot post-filter")
  vln_list <- lapply(c("nFeature_RNA", "nCount_RNA", "percent.mt"), function(f) {
    VlnPlot(srt_filtered, features = f, pt.size = 0.3, alpha = 0.1) +
    scale_y_continuous(breaks = pretty_breaks(n = 7)) +
    theme(legend.position = "none") 
  })
  
  print("combining violin plots pre-filter")
  post_vln <- wrap_plots(vln_list, ncol = 3) +
  plot_annotation(title = paste0("Post-filter: ", sample_name))
  
  # --- Combine and save the plots ---
  print("combining violin plots pre and post-filter")
  combined_vln <- pre_vln / post_vln  # Vertical stacking with patchwork
  ggsave(filename = paste0(output_dir, "qc_violin_", sample_name, ".jpg"),
         plot = combined_vln, width = 12, height = 10)
  print(combined_vln)

  # --- Scatterplots before filtering ---
  print("create FeatureScatter plots")
  p1 <- FeatureScatter(srt_obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
  p2 <- FeatureScatter(srt_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
  plot_combined <- p1 + p2
  ggsave(filename = paste0(output_dir, "featurescatter_prefilter_", sample_name, ".jpg"),
         plot = plot_combined, width = 12, height = 6)
  
  message("QC complete for: ", sample_name)
  return(srt_filtered)
}
```

# Define sample names
```{r}
sample_names <- c("P1_Pre",
                  "P1_Post",
                  "P4_Pre",
                  "P4_Post",
                  "P7_Pre",
                  "P7_Post",
                  "P13_Pre",
                  "P13_Post",
                  "P14_Post",
                  "P16_Post")
```

# First import post-cellbender counts
```{r}
# Base directory where cellbender counts are stored
base_dir <- "~/processed_data/cellbender"

#Initialize empty list to store seurat objects
seurat_objects_cellbender <- list()

# Loop through each sample name and process it
for (sample in sample_names) {
  
  # Construct the full path dynamically
  sample_path <- file.path(base_dir, sample, "cellbender_output_FPR_0.0_filtered_seurat.h5")

  # Check if the file exists before proceeding
  if (file.exists(sample_path)) {
    print(sample_path)
    # now import and construct seurat object
    seurat_obj <- CreateSeuratObject(counts = Read10X_h5(sample_path),
                                     project = sample)
    seurat_objects_cellbender[[sample]] <- seurat_obj
  } else {
    warning(paste("File not found:", sample_path))
  }
}

# Check the list of Seurat objects
seurat_objects_cellbender

```

# get number of cells for each sample pre-filter
```{r}
nCells_pre_filter <- lapply(seurat_objects_cellbender, ncol)
names(nCells_pre_filter) <- names(seurat_objects_cellbender)
nCells_pre_filter
```


# Import scrublet doublet calls and filter cellbender output to exlude doublets
```{r}
base_dir <- "~/processed_data/scrublet/cell_resubmission/"
seurat_objects_scrublet <- list()

for(sample in sample_names){
  scrublet_dir <- file.path(base_dir, sample, "scrublet_doublets.csv")
  
  scrublet_df <- read_csv(scrublet_dir) %>%
    dplyr::rename("cell_id" = "...1")
  
  predicted_doublets <- sum(scrublet_df$predicted_doublet[!is.na(scrublet_df$predicted_doublet)])
  
  print(paste0("number of scrublet-predicted doublets: ", predicted_doublets))
  
  non_doublets <- scrublet_df %>% dplyr::filter(predicted_doublet == FALSE) %>% 
    dplyr::pull(cell_id)
  
  seurat_objects_scrublet[[sample]] <- subset(seurat_objects_cellbender[[sample]], cells = non_doublets)
  }
```


# import filtered cell matrices from cell ranger
these will be cross checked with the cellbender output
```{r}
# Base directory where the raw data is stored
base_dir <- "~/RawData"

# Initialize an empty list to store Seurat objects
seurat_objects_cellranger <- list()

# Loop through each sample name and process it
for (sample in sample_names) {
  
  # Construct the full path dynamically
  sample_path <- file.path(base_dir, sample, sample)
  
  # Construct the path to the H5 file
  h5_path <- file.path(sample_path, "filtered_feature_bc_matrix.h5")
  
  # Check if the file exists before proceeding
  if (file.exists(h5_path)) {
    # Load the counts matrix
    counts <- Read10X_h5(h5_path)
    # Create a Seurat object
    seurat_obj <- CreateSeuratObject(counts = counts, 
                                     project = sample)
    # Store the Seurat object in the list
    seurat_objects_cellranger[[sample]] <- seurat_obj
    # Print a message indicating successful processing
    print(paste("Processed:", sample))
    
  } else {
    warning(paste("File not found:", h5_path))
  }
}

# Check the list of Seurat objects
seurat_objects_cellranger
```

# filter cellbender-processed seurat objects to cells also in the processed cellranger output
```{r}
seurat_objects_consensus <- list()
for(sample in sample_names){
  print(sample)
  cr_cells <- colnames(seurat_objects_cellranger[[sample]])
  seurat_objects_consensus[[sample]] <- subset(seurat_objects_scrublet[[sample]], cells = cr_cells)
  }
```


# filter seurat objects manually
```{r}
seurat_qc <- list()
params_df <- read_csv("path/to/resources/qc_filtering_params_strong_revisions.csv") # you have to make this file and input filtering parameters yourself
sample_names

for(sample in sample_names){
  params <- params_df[params_df$sample == sample, ]

  min_feat <- params$min_features
  max_feat <- params$max_features
  min_cts  <- params$min_counts
  max_cts  <- params$max_counts

  seurat_qc[[sample]] <- qc_filter_seurat(seurat_objects_consensus[[sample]], sample_name = sample,
                                          min_features = min_feat, max_features = max_feat, 
                                          min_counts = min_cts, max_counts = max_cts)
  
}
```

# summarize QC stats
```{r}
cb_counts <- c()
qc_counts <- c()

for(sample in sample_names){
  cb_counts <- c(cb_counts, ncol(seurat_objects_cellbender[[sample]]))
  qc_counts <- c(qc_counts, ncol(seurat_qc[[sample]]))
}

qc_filtering_params_stats_tumor <- tibble("sample" = sample_names,
                                              "tissue_type" = "tumor",
                                              "nCells_prefilter" = cb_counts,
                                              "nCells_postfilter" = qc_counts,
                                              "fraction_filtered" = ((cb_counts - qc_counts)/cb_counts)
)
params_df$max_percent_mitochondrial <- 10
qc_filtering_params_stats_tumor <- left_join(x = params_df, qc_filtering_params_stats_tumor)

```

```{r}
# save qc stats
write_csv(qc_filtering_params_stats_tumor, file = "~/results/qc_filtering_params_results_stats_tumor_revisions.csv")
```


# normalize each seurat object. It is necessary to normalize each tumor separately.
```{r}
# Initialize a list to store fully processed Seurat objects
seurat_normalized <- list()
# Loop through each Seurat object to process it
for (sample in names(seurat_qc)) {
  print(sample)
  # Retrieve the QC-filtered Seurat object
  srt_obj <- seurat_qc[[sample]]
  
  # Normalize data
  srt_obj <- NormalizeData(srt_obj, normalization.method = "LogNormalize", scale.factor = 10000)
  seurat_normalized[[sample]] <- srt_obj
  # Print success message
  print(paste("Normalization completed for:", sample))
}

# Check the final list of processed Seurat objects
```

# remove unwanted seurat objects to clear memory
```{r}
rm(seurat_objects_cellbender)
rm(seurat_objects_cellranger)
rm(seurat_objects_consensus)
rm(seurat_objects_scrublet)
rm(seurat_qc)
gc()
```


# Merge seurat objects and export merged obj.
```{r}
processed_data_dir <- "~/processed_data/"

# append sample id to cell names. this avoids overlapping cell barcodes between samples
for(sample in sample_names){
  colnames(seurat_normalized[[sample]]) <- paste(colnames(seurat_normalized[[sample]]), seurat_normalized[[sample]]$orig.ident, sep = "_")
}

# merge
srt_merge <- merge(
  x = seurat_normalized[[1]],
  y = seurat_normalized[2:length(seurat_normalized)],
  merge.data = TRUE
)

# In Seurat v5 you have to manually join all the layers.
# separately join normalized layers (data) and raw counts "counts"
counts_layers <- grep("^counts", Layers(srt_merge), value = TRUE)
norm_layers <- grep("^data", Layers(srt_merge), value = TRUE)
counts_layers
norm_layers

srt_merge <- JoinLayers(srt_merge)
Layers(srt_merge)

```


# Find variable features and scale merged dataset
It is necessary to find variable features and scale after merging as the merge alters these values
```{r}
markers_df <- read_csv("~/resources/cell_type_markers_all.csv")
genes_to_scale <- markers_df$marker

processed_data_dir <- "~/processed_data/"

# Find variable features
srt_merge <- FindVariableFeatures(srt_merge, selection.method = "vst", nfeatures = 2000)

# Scale data - R crashes if you try to scale with all features. So only scale variable features and genes_to_scale for later annotation.
srt_merge <- ScaleData(srt_merge, features = c(VariableFeatures(srt_merge), genes_to_scale)) 
```

# Add other key metadata
```{r}
# basic sample info
patient <- srt_merge@meta.data %>%
  pull(orig.ident) %>%
  word(start = 1, end = 1, sep = fixed("_"))

timepoint <- srt_merge@meta.data %>%
  pull(orig.ident) %>%
  word(start = 2, end = 2, sep = fixed("_"))

sample_name <- srt_merge@meta.data %>%
  pull(orig.ident) %>%
  word(start = 1, end = 2, sep = fixed("_"))

# response status
responders <- c("P4", "P7", "P13", "P16")
nonresponders <- c("P1", "P14")

response_status <- srt_merge@meta.data %>%
  mutate(response_status = case_when(
    patient %in% responders ~ "responder",
    patient %in% nonresponders ~ "nonresponder"
  )) %>%
  pull(response_status)
response_status %>% table()

# add columns to metadata
srt_merge <- AddMetaData(srt_merge, metadata = patient, col.name = "patient")
srt_merge <- AddMetaData(srt_merge, metadata = timepoint, col.name = "timepoint")
srt_merge <- AddMetaData(srt_merge, metadata = sample_name, col.name = "sample_name")
srt_merge <- AddMetaData(srt_merge, metadata = response_status, col.name = "response_status")

# check columns
srt_merge$patient %>% table()
srt_merge$timepoint %>% table()
srt_merge$sample_name %>% table()
srt_merge$response_status %>% table()
```

```{r}
# check metadata
meta <- srt_merge@meta.data
```


```{r}
# save seurat obj for use in next script
saveRDS(srt_merge, file = paste0("/path/to/tumors_normalized_merged_scaled_srt_revision.rds"))
```

