---
title: "tumor_analysis_inferCNV_P16_Post"
output: html_document
date: "2025-12-12"
---

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(readr)
library(infercnv)
```


# Define functions
This outputs annotations files for cell ranger cell ids and seurat clusters and returns a gene expression matrix
```{r}
# Define function to process each Seurat object with user-defined clusters
process_seurat_annotations <- function(seurat_obj, sample_name,
                                       output_dir = "~/resources/") {
  
  # Ensure output directory exists
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Extract gene expression matrix
  GEX_matrix <- as.matrix(GetAssayData(seurat_obj, assay = "RNA", slot = "counts"))
  
  # Create annotations for Cell Ranger and Seurat clusters
  annotations_seurat <- as_tibble(FetchData(seurat_obj, vars = "seurat_clusters"), rownames = "cellID")
  annotations_inferCNV <- as_tibble(FetchData(seurat_obj, vars = "infercnv_annotation"), rownames = "cellID")

  # Save annotation files without column names
  write_tsv(annotations_seurat, paste0(output_dir, "annotations_seurat_inferCNV_figure_", sample_name, ".tsv"), col_names = FALSE)
  write_tsv(annotations_inferCNV, paste0(output_dir, "annotations_sample_vs_control_inferCNV_figure_", sample_name, ".tsv"), col_names = FALSE)

  # Print completion message
  print(paste("Gene expression and annotation files saved for:", sample_name))
  
  if(all(colnames(GEX_matrix) %in% annotations_seurat$cellID)){
    print("all cells in matrix are in the annotation data")
  }
  # Return processed data as a list (optional)
  return(GEX_matrix)
}

```



```{r}
run_inferCNV <- function(gene_expression_matrix, annotation_file, gene_order_file, output_dir = "~/results/inferCNV/", ref_groups) {
  
  # Create InferCNV object
  print("create infercnv object")
  inferCNVObj <- CreateInfercnvObject(
    raw_counts_matrix = gene_expression_matrix,
    gene_order_file = gene_order_file,
    annotations_file = annotation_file,
    delim = '\t',
    ref_group_names = ref_groups
  )
  
  # Run InferCNV
  print("run inferCNV")
  inferCNV_res <- infercnv::run(
    inferCNVObj, 
    cutoff = 1, 
    out_dir = output_dir,
    cluster_by_groups = TRUE,
    denoise = TRUE,
    HMM = FALSE,  # Set to TRUE if using Hidden Markov Model for CNV calling
    useRaster = FALSE,
    png_res = 900,
    #output_format = "pdf"
    output_format = "png",
    num_threads = 12
    )
  
  # Print completion message
  print(paste("InferCNV completed and results saved in:", output_dir))
  
  # Return the InferCNV result object
  return(inferCNV_res)
}

```

```{r}
set.seed(1234)
```


# import P16 Post from script 4
```{r}
srt_P16_Post <- read_rds("~/processed_data/tumors_normalized_merged_scaled_annotated_coarse_srt_revision_P16_Post.rds")
```

```{r}
# check cell type distribution
DimPlot(srt_P16_Post, group.by = "cell_type_manual")
DimPlot(srt_P16_Post, group.by = "seurat_clusters")
srt_P16_Post$cell_type_manual %>% table()
```


```{r}
# Read malignant cluster calls
malig_file <- "/resources/malignant_cell_calls_tumor_P16_Post_revision.txt"
malig_clusters <- read_lines(malig_file)

# Get number of clusters
nClusters <- length(table(FetchData(srt_P16_Post, vars = "seurat_clusters")))
  
# Determine non-malignant clusters
nonmalig_clusters <- as.character(setdiff(0:(nClusters - 1), malig_clusters))

malig_clusters
nonmalig_clusters

```

# create infercnv annotations in metadata to separate by malignant vs. nonmalignant
This creates the "inferCNV annotation" column which is needed for "process_seurat_annotations"
```{r}
# create a new metadata column that labels cells as either sample name or Nonmalignant
srt_P16_Post@meta.data <- srt_P16_Post@meta.data %>% 
  mutate(malignant = case_when(seurat_clusters %in% malig_clusters ~ "malignant",
                               TRUE ~ "nonmalignant")) 
srt_P16_Post@meta.data$malignant %>% table()

# divide nonmalignant into "glia", "lymphoid", "myeloid", and "other
srt_P16_Post$cell_type_manual %>% table()
srt_P16_Post$sample_name %>% table()
meta <- srt_P16_Post@meta.data

infercnv_annotation <- meta %>%
  mutate(infercnv_annotation = case_when(
    malignant == "malignant" ~ sample_name,
    TRUE ~ "Other"
    )) %>%
  dplyr::pull(infercnv_annotation)

# check that the vector looks right
infercnv_annotation %>% table()

# add to metadata - this is key for creating the "by_sample" annotation column
srt_P16_Post <- AddMetaData(srt_P16_Post, infercnv_annotation, col.name = "infercnv_annotation")

srt_P16_Post$infercnv_annotation %>% table()
srt_P16_Post$cell_type_manual %>% table()
srt_P16_Post$seurat_clusters %>% table()
dim(srt_P16_Post)
```


# create annotation files and GEX matrix for each sample
```{r}
out_dir <- "~/resources/"

# Process annotations
GEX_matrix <- process_seurat_annotations(
  seurat_obj = srt_P16_Post,
  sample_name = "tumors_P16_Post_revision", # this is different from the sample name input into run infercnv. This changes the suffix of the annotation files
  output_dir = out_dir)
```

```{r}
#By Seurat cluster

# set up resources 
annotations_file <- "~/resources/annotations_seurat_inferCNV_figure_tumors_P16_Post_revision.tsv"
out_dir_base <- "~/results/inferCNV/"
gof <- "~/external_resources/hg38_gencode_v27.txt"

sample_name <- "tumors_by_cluster_figure_revision_P16_Post"


# important diagnostic! Check that all annotations cells are in the GEX matrix!
annotations <- read_tsv(annotations_file, col_names = FALSE)
all(colnames(GEX_matrix) %in% annotations$X1)

ref_groups <- nonmalig_clusters
annotations$X2 %>% table()

res <- run_inferCNV(gene_expression_matrix = GEX_matrix,
             annotation_file = annotations_file,
             gene_order_file = gof,
             output_dir = paste0(out_dir_base, sample_name),
             ref_groups = ref_groups)
```

# Find the barcodes of bona fide P16 post malignant cells from inferCNV result object
```{}
P16 post has only about 11 malignant cells (visualized by infeCNV output) in cluster 11. These cells do not cluster separately from nonmalignant cells. It is therefore necessary to ID these few cells by barcode and manually update their cell type to malignant in the main seurat object containing all tumor cells.
```


```{r}
# get hierarchical clustering data
hc <- res@tumor_subclusters$hc

# get barcodes in the same order as in the dendrogram.
cell_order <- hc$'11'$labels[hc$'11'$order]

cell_order

df <- data.frame(
  cell_id = cell_order,
  dendrogram_position = seq_along(cell_order)
)

# get index of the firs and last malignant cell, which all sub-clustered together in inferCNV
# last malignant cell is 2 from the end. First malignant cell is 10 before that. 11 total.
#first_malig <- dim(df)[1] - 2 - 10
#last_malig <- dim(df)[1] - 2

# pull malignant cells
#malignant_cells <- df$cell_id[first_malig:last_malig]
malignant_cells <- df$cell_id[3:13]
malignant_cells
```

# update P16 seurat object
```{r}
# Identify cells in cluster 11
cells_cluster11 <- WhichCells(
  object = srt_P16_Post,
  idents = 11
)

# check this is the right cluster
cells_cluster11_test <- srt_P16_Post@meta.data %>%
  filter(seurat_clusters == "11") %>%
  rownames()

all(cells_cluster11_test == cells_cluster11)

# Assign all cluster 11 cells as ambiguous.
srt_P16_Post@meta.data[cells_cluster11, "cell_type_manual"] <- "Ambiguous"
srt_P16_Post@meta.data[cells_cluster11, "malignant"] <- "nonmalignant"

srt_P16_Post$malignant %>% table()
srt_P16_Post$cell_type_manual %>% table()

srt_P16_Post@meta.data %>% filter(seurat_clusters == "11")

# Assign called malignant cells as "malignant" = "malignant and "cell_type_manual" = "malignant
srt_P16_Post@meta.data[malignant_cells, "cell_type_manual"] <- "Malignant"
srt_P16_Post@meta.data[malignant_cells, "malignant"] <- "malignant"

srt_P16_Post$malignant %>% table()
```

# Test bona fide malignant cells with inferCNV.
Set up first
```{r}

# divide nonmalignant into "glia", "lymphoid", "myeloid", and "other
srt_P16_Post$cell_type_manual %>% table()
srt_P16_Post$sample_name %>% table()

meta <- srt_P16_Post@meta.data

infercnv_annotation <- meta %>%
  mutate(infercnv_annotation = case_when(
    malignant == "malignant" ~ sample_name,
    malignant == "nonmalignant" & cell_type_manual == "Ambiguous" ~ "Ambiguous",
    TRUE ~ "Other"
    )) %>%
  dplyr::pull(infercnv_annotation)

# check that the vector looks right
infercnv_annotation %>% table()

# add to metadata - this is key for creating the "by_sample" annotation column
srt_P16_Post <- AddMetaData(srt_P16_Post, infercnv_annotation, col.name = "infercnv_annotation")

srt_P16_Post$infercnv_annotation %>% table()
srt_P16_Post$cell_type_manual %>% table()
srt_P16_Post$seurat_clusters %>% table()
dim(srt_P16_Post)
```


# create annotation files and GEX matrix for each sample
```{r}
out_dir <- "~/resources/"

# Process annotations
GEX_matrix <- process_seurat_annotations(
  seurat_obj = srt_P16_Post,
  sample_name = "tumors_P16_Post_revision_refined", # this is different from the sample name input into run infercnv. This changes the suffix of the annotation files
  output_dir = out_dir)
```


# run InferCNV with updated malignant cells.

```{r}
# set up resources
annotations_file <- "~/resources/annotations_sample_vs_control_inferCNV_figure_tumors_P16_Post_revision_refined.tsv"
out_dir_base <- "~/results/inferCNV/"
gof <- "~/external_resources/hg38_gencode_v27.txt"


sample_name <- "tumors_merge_by_sample_figure_revision_P16_Post_Refined"

# important diagnostic! Check that all annotations cells are in the GEX matrix!
annotations <- read_tsv(annotations_file, col_names = FALSE)
all(colnames(GEX_matrix) %in% annotations$X1)
ref_groups <- c("Ambiguous", "Other")

annotations$X2 %>% table()

run_inferCNV(gene_expression_matrix = GEX_matrix,
             annotation_file = annotations_file,
             gene_order_file = gof,
             output_dir = paste0(out_dir_base, sample_name),
             ref_groups = ref_groups)
```


```{r}
srt_merge <- readRDS("~/processed_data/tumors_normalized_merged_scaled_annotated_coarse_srt_revision.rds")
srt_merge$sample_name %>% table()
srt_merge$cell_type_manual %>% table()
```

```{r}
P16_Post_cells_called_Malig <- srt_merge@meta.data %>%
  dplyr::filter(sample_name == "P16_Post") %>%
  dplyr::filter(cell_type_manual == "Malignant") %>%
  rownames()

length(P16_Post_cells_called_Malig)

# overwrite incorrect malignant cells to ambiguous.
srt_merge@meta.data[P16_Post_cells_called_Malig, "cell_type_manual"] <- "Ambiguous"

# overwrite bona fide malignant cells accordingly
srt_merge@meta.data[malignant_cells, "cell_type_manual"] <- "Malignant"
```

```{r}
saveRDS(srt_merge, "~/processed_data/tumors_normalized_merged_scaled_annotated_coarse_malig_fixed_srt_revision.rds")
```

```{r}
srt_merge <- readRDS("~/processed_data/tumors_normalized_merged_scaled_annotated_coarse_malig_fixed_srt_revision.rds")
```


```{r}
# check malignant count. P16 post should have 11 malig.
malig_count <- srt_merge@meta.data %>%
  subset(cell_type_manual == "Malignant") %>%
  group_by(sample_name) %>%
  summarize(malig_count = n())
```


# subset seurat obj

```{r}
set.seed(1234)
```

```{r}
# InferCNV cannot handle all ~ 100000 cells in the dataset. so we will subset the same proportion of cells for each cluster
meta <- srt_merge@meta.data
meta$cell <- rownames(meta)

target_n <- 15000
sample_proportion <- target_n/nrow(meta)

sampled_cells <- meta %>% # take metadata
  group_by(seurat_clusters) %>%   # group by samples
  slice_sample(prop = sample_proportion) %>% # take a random proportion of each cluster
  pull(cell)

srt_merge_sub <- subset(srt_merge, cells = sampled_cells)
srt_merge_sub
```


```{r}
# clear big seurat obj from memory
rm(srt_merge)
gc()
```


```{r}
# create a new metadata column that labels cells as either sample name or Nonmalignant
srt_merge_sub@meta.data <- srt_merge_sub@meta.data %>% 
  mutate(Malignant = case_when(cell_type_manual == "Malignant" ~ "Malignant",
                               TRUE ~ "Nonmalignant")) 

srt_merge_sub@meta.data$Malignant %>% table()
srt_merge_sub$cell_type_manual %>% table()
# divide nonmalignant into "glia", "lymphoid", "myeloid", and "other
meta <- srt_merge_sub@meta.data

infercnv_annotation <- meta %>%
  mutate(infercnv_annotation = case_when(
    Malignant == "Nonmalignant" & cell_type_manual == "Glial cell" ~ "Glial cell",
    Malignant == "Nonmalignant" & cell_type_manual == "Lymphoid" ~ "Lymphoid",
    Malignant == "Nonmalignant" & cell_type_manual == "Myeloid" ~ "Myeloid",
    Malignant == "Malignant" ~ sample_name,
    TRUE ~ "Other"
    )) %>%
  dplyr::pull(infercnv_annotation)

# check that the vector looks right
infercnv_annotation %>% table()

# add to metadata
srt_merge_sub <- AddMetaData(srt_merge_sub, infercnv_annotation, col.name = "infercnv_annotation")

srt_merge_sub$infercnv_annotation %>% table()
srt_merge_sub$seurat_clusters %>% table()
dim(srt_merge_sub)
```

# create annotation files and GEX matrix for each sample
```{r}
#out_dir <- "~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/resources/"
out_dir <- "~/resources/"

# Process annotations
GEX_matrix <- process_seurat_annotations(
  seurat_obj = srt_merge_sub,
  sample_name = "tumors_revision_malig_fixed", # this is different from the sample name input into run infercnv. This changes the suffix of the annotation files
  output_dir = out_dir)
```

# run InferCNV with fixed malignant cells
cluster by sample
```{r}
# set up resources
annotations_file <- "~/resources/annotations_sample_vs_control_inferCNV_figure_tumors_revision_malig_fixed.tsv"
out_dir_base <- "~/results/inferCNV/"
gof <- "~/external_resources/hg38_gencode_v27.txt"

sample_name <- "tumors_merge_by_sample_figure_revision_malig_fixed"

# important diagnostic! Check that all annotations cells are in the GEX matrix!
annotations <- read_tsv(annotations_file, col_names = FALSE)
all(colnames(GEX_matrix) %in% annotations$X1)
ref_groups <- c("Lymphoid", "Myeloid", "Glial cell", "Other")
annotations$X2 %>% table()

res <- run_inferCNV(gene_expression_matrix = GEX_matrix,
             annotation_file = annotations_file,
             gene_order_file = gof,
             output_dir = paste0(out_dir_base, sample_name),
             ref_groups = ref_groups)
```



