---
title: "tumor_preprocessing_8_manual_annotations_miller_myeloid.rmd"
output: html_document
date: "2025-12-18"
---


```{r}
library(Seurat)
library(readr)
library(tibble)
library(dplyr)
library(viridis)
library(presto) # used to greatly improve FindAllMarkers efficiency
```

# define funcitons
```{r}
# Define a function to run UMAP
create_UMAP <- function(seurat_obj, sample_name, num_pcs, res = 0.25) {

  # Find neighbors and clusters
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs)
  seurat_obj <- FindClusters(seurat_obj, resolution = res)
  
  # Print cluster assignments for verification
  print(paste("Clusters for", sample_name, ":"))
  print(head(Idents(seurat_obj)))
  
  # Run UMAP
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs)

  # Save UMAP plots with different groupings
  #p1 <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + ggtitle(sample_name)
  #print(p1)

  # Print message indicating completion
  print(paste("UMAP and clustering completed for:", sample_name))

  # Return the updated Seurat object
  return(seurat_obj)
}
```

# Import tumor cells with coarse annotations
```{r}
srt_merge <- readRDS("~/processed_data/tumors_normalized_scaled_dimreduction_annotated_refined_srt_revision.rds")
srt_merge$cell_type_manual %>% table()
```

```{r}
# load miller lab myeloid annotations based on their existing programs
miller_df <- read.csv("/resources/edward_penn-silverbush-cart-collaboration-2025_ege-miller-silverbush-2025-slvb_myeloid_bc_usages_final_annotated.csv") 
colnames(miller_df)[1] <- "cell_id"
```


The miller lab annotations have 492 cells that are not cell types but are instead just activities. All cells that have program-related annotations have no cell type annotation. The programs add up to 492 which is the same as the number of cells without types. For these, it is necessary to fall back to my manual annotations.

Also: Use "cell_identity_annotation" not "cell_type_identity_annotation" as 

```{r}
miller_df$cell_identity_annotation %>% table(useNA = "ifany")
table(miller_df$cell_identity_annotation == "", useNA = "ifany")
miller_df$final_annotation %>% table()

```


# update seurat object with miller lab myeloid cell annotations
```{r}
# extract cell barcodes in both miller lab annotations and main seurat object (myeloid cells)
matching_cells <- colnames(srt_merge)[colnames(srt_merge) %in% miller_df$cell_id]

# get index of common (myeloid) cells in miller annotations
match_index <- match(matching_cells, miller_df$cell_id)

# initialize a final annotation column
srt_merge$cell_type_final <- srt_merge$cell_type_manual_refined
srt_merge$cell_type_final_no_program <- srt_merge$cell_type_manual_refined

# overwrite myeloid cell annotations with miller lab annotations
srt_merge$cell_type_final[matching_cells] <- miller_df$final_annotation[match_index]

# only overwrite annotations for cells with valid cell type annotations

# pull Miller annotations aligned to matching_cells, correctly ordered
miller_no_program <- miller_df$cell_identity_annotation[match_index]
# identify valid (non-empty, non-NA) entries
valid <- !is.na(miller_no_program) & miller_no_program != ""
# overwrite ONLY those
srt_merge$cell_type_final_no_program[matching_cells[valid]] <- miller_no_program[valid]
```

```{r}
srt_myeloid <- subset(srt_merge, subset = cell_type_manual == "Myeloid")

# check if all miller lab annotation cells are in srt_myeloid and vice versa
all(colnames(srt_myeloid) %in% miller_df$cell_id) & all(miller_df$cell_id %in% colnames(srt_myeloid)) # Returns TRUE

```

```{r}
srt_myeloid$cell_type_final_no_program %>% table()
```


# re-cluster and check umap
```{r}
set.seed(1234)

# find variable features
srt_myeloid <- FindVariableFeatures(srt_myeloid, selection.method = "vst", nfeatures = 2000)

# Scale data 
markers_df <- read_csv("~/resources/Myeloid_markers_revision.csv")

srt_myeloid <- ScaleData(srt_myeloid, features = VariableFeatures(srt_myeloid))

srt_myeloid <- RunPCA(srt_myeloid, features = VariableFeatures(object = srt_myeloid))

plot <- ElbowPlot(srt_myeloid)
plot
```

From elbow plot, use 10 PCs
```{r}
sample_name <- "myeloid_refined_revision"
srt_myeloid <- create_UMAP(srt_myeloid, sample_name = sample_name, 
        num_pcs = 10, res = 0.35)

srt_myeloid@meta.data %>% head()
```

```{r}
# and plot over UMAP to be sure:
p1 <- DimPlot(srt_myeloid, group.by = "cell_type_manual_refined")
p2 <- DimPlot(srt_myeloid, group.by = "cell_type_final")
p3 <- DimPlot(srt_myeloid, group.by = "cell_type_final_no_program")

ggsave("~/results/dimreduction/umap/umap_tumor_myeloid_annotation_miller_full_assignments_revision.png", width = 8, height = 5, plot = p2)
ggsave("~/results/dimreduction/umap/umap_tumor_myeloid_annotation_miller_type_only_assignments_revision.png", width = 8, height = 5, plot = p3)

p1
p2
p3
```

```{r fig.width=10, fig.height=5}
DimPlot(srt_merge, group.by = "cell_type_manual_refined")
```


```{r}
colnames(srt_merge@meta.data)
```

```{r fig.width= 10, fig.height= 5}
DimPlot(srt_merge, group.by = "cell_type_final_no_program")
```


```{r}
saveRDS(srt_merge, file = "~/processed_data/tumors_normalized_scaled_dimreduction_annotated_refined_miller_srt_revision.rds")
```


