---
title: "E20_Script6_Manual_Annotation"
output: html_document
date: "2025-04-01"
---

```{r}
# for presto:
#devtools::install_github("immunogenomics/presto")
```


```{r}
library(Seurat)
library(readr)
library(tibble)
library(dplyr)
library(viridis)
library(presto) # used to greatly improve FindAllMarkers efficiency
```

# define funcitons
```{r}
# Define a function to run UMAP
create_UMAP <- function(seurat_obj, sample_name, num_pcs, res = 0.25) {

  # Find neighbors and clusters
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs)
  seurat_obj <- FindClusters(seurat_obj, resolution = res)
  
  # Print cluster assignments for verification
  print(paste("Clusters for", sample_name, ":"))
  print(head(Idents(seurat_obj)))
  
  # Run UMAP
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs)

  # Save UMAP plots with different groupings
  #p1 <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + ggtitle(sample_name)
  #print(p1)

  # Print message indicating completion
  print(paste("UMAP and clustering completed for:", sample_name))

  # Return the updated Seurat object
  return(seurat_obj)
}
```

# function to plot markers
```{r}
plot_marker_umaps <- function(seurat_obj, marker_csv, samplename = "sample", path_out) {
  # Load marker gene list
  marker_df <- read_csv(marker_csv)
  
  # Validate columns
  if (!all(c("cell_type", "marker") %in% colnames(marker_df))) {
    stop("CSV must contain 'cell_type' and 'marker' columns")
  }
  
  # Drop NA or blank marker entries
  marker_df <- marker_df[!is.na(marker_df$marker) & marker_df$marker != "", ]
  
  plot_list <- list()

  # First: Add UMAP with Seurat cluster identity
  cluster_plot <- DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5) +
    ggtitle("Clusters")
  plot_list[[1]] <- cluster_plot

  # Then: Add FeaturePlots for each marker
  for (i in seq_len(nrow(marker_df))) {
    marker <- marker_df$marker[i]
    cell_type <- marker_df$cell_type[i]
    
    if (marker %in% rownames(seurat_obj)) {
      p <- FeaturePlot(seurat_obj, features = marker, reduction = "umap", pt.size = 0.25, order = TRUE) +
        ggtitle(paste0(cell_type, ": ", marker))
    } else {
      p <- ggplot() + 
        theme_void() + 
        ggtitle(paste0(cell_type, ": ", marker, " (not found)"))
    }
    plot_list[[i + 1]] <- p  # shift index by 1 to account for cluster plot
  }
  
  # Layout: adjust ncol to control grid width
  ncol <- 6
  nrow <- ceiling(length(plot_list) / ncol)
  final_plot <- wrap_plots(plotlist = plot_list, ncol = ncol)
  
  # Save
  jpg_filename <- paste0(path_out, '/', samplename, ".jpg")
  ggsave(jpg_filename, final_plot, width = 4 * ncol, height = 3 * nrow)
  
  message("Saved plot to: ", jpg_filename)
  return(final_plot)
}
```

```{r}
set.seed(1234)
```

#import filtered/preprocessed seurat objects
```{r}
srt_merge <- readRDS("~/processed_data/tumors_normalized_merged_scaled_dimreduction_srt_revision.rds")
```

```{r}
markers_path <- "~/resources/cell_type_markers_for_multi_marker_UMAP.csv" # this is a file containing the markers and associated cell type. Change only if you want new markers.

plot_marker_umaps(srt_merge, markers_path, samplename = "UMAP_annotation_tumors_revisions_test", path_out = "~/results")
```


# import manual cluster-cell type annotations
```{r}
annotations.df <- read_csv("~/resources/manual_annotations/annotation_tumors_all_revision.csv")

clusters <- srt_merge$seurat_clusters
celltypes <- annotations.df$cell_type[match(clusters, annotations.df$seurat_cluster)] # use the match function to add manually called cell types for each cluster to each cell.
srt_merge <- AddMetaData(srt_merge, celltypes, col.name = "cell_type_manual")

srt_merge@meta.data$cell_type_manual %>% table()
srt_merge
DimPlot(srt_merge, group.by = "cell_type_manual")
```

# export to inferCNV before proceeding to validate malignant cell calls.
takes a long time. run only when necessary.
```{r}

# saveRDS(srt_merge, "~/processed_data/tumors_normalized_merged_scaled_annotated_coarse_srt_revision.rds")
```

```{}
Some malig cell calls are spurious in P16 post. Subcluster P16 and re-assess.
```

# subset to P16 Post

```{r}
srt_P16_Post <- readRDS("~/processed_data/tumors_normalized_merged_scaled_annotated_coarse_srt_revision_P16_Post.rds")
```

```{r}
srt_P16_Post <- subset(srt_merge, subset = sample_name == "P16_Post")
# check cell type distribution
DimPlot(srt_P16_Post, group.by = "cell_type_manual")
srt_P16_Post$cell_type_manual %>% table()
```

# Find variable features and scale merged dataset
It is necessary to find variable features and scale after merging as the merge alters these values
```{r}
markers_df <- read_csv("~/resources/cell_type_markers_all.csv")
genes_to_scale <- markers_df$marker

processed_data_dir <- "~/processed_data/"

# Find variable features
srt_P16_Post <- FindVariableFeatures(srt_P16_Post, selection.method = "vst", nfeatures = 2000)

# Scale data - R crashes if you try to scale with all features. So only scale variable features and genes_to_scale for later annotation.
srt_P16_Post <- ScaleData(srt_P16_Post, features = c(VariableFeatures(srt_P16_Post), genes_to_scale), vars.to.regress = c("nCount_RNA", "nFeature_RNA")) 
```


```{r}
# Define results directory
dimreduction_dir <- "~/results/dimreduction/pca/"
sample <- "tumor_filtered_processed_annotated_P16_post"

# Run PCA
srt_P16_Post <- RunPCA(srt_P16_Post, features = VariableFeatures(object = srt_P16_Post))

# Save ElbowPlot
p4 <- ElbowPlot(srt_P16_Post)
print(p4)
```

From the elbowplot, ideal number of dimensions is:
7
```{r}
# use the function for each object
srt_P16_Post <- create_UMAP(srt_P16_Post, "merged_filtered_processed_revision_P16_Post", res = 0.5, num_pcs =  6)
```


```{r}
p1 <- DimPlot(srt_P16_Post, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
#ggsave(filename = "~/results/dimreduction/umap/UMAP_merged_filtered_processed_revision_seurat_clusters_P16_Post.jpg",
         plot = p1, dpi = 150, height = 10, width = 10)
p1
```

```{r}
markers_path <- "~/resources/cell_type_markers_for_multi_marker_UMAP.csv" # this is a file containing the markers and associated cell type. Change only if you want new markers.
plot_marker_umaps(srt_P16_Post, markers_path, samplename = "UMAP_annotation_tumors_revisions_P16_Post", path_out = "~/results")
```

```{r}
meta <- srt_P16_Post@meta.data
FeaturePlot(srt_P16_Post, features = "EGFR")
FeaturePlot(srt_P16_Post, features = "nCount_RNA")
FeaturePlot(srt_P16_Post, features = "nFeature_RNA")

```

# import manual cluster-cell type annotations - P16 Post
```{r}
annotations.df <- read_csv(paste0("~/resources/annotation_tumors_P16_post_revision.csv"))

clusters <- srt_P16_Post$seurat_clusters
celltypes <- annotations.df$cell_type[match(clusters, annotations.df$seurat_cluster)] # use the match function to add manually called cell types for each cluster to each cell.
srt_P16_Post <- AddMetaData(srt_P16_Post, celltypes, col.name = "cell_type_manual")

srt_P16_Post@meta.data$cell_type_manual %>% table()
```

```{r}
markers_11 <- FindMarkers(srt_P16_Post, ident.1 = "11") %>%
  rownames_to_column(var = "geneID")

write_tsv(markers_11, "~/results/P16_post_markers_clust_11.tsv")
```


```{r}
# export to validate in inferCNV
saveRDS(srt_P16_Post, file = "~/processed_data/tumors_normalized_merged_scaled_annotated_coarse_srt_revision_P16_Post.rds")
```

