---
title: "Figure3"
output: html_document
date: "2025-04-18"
---


```{r}
install.packages("ggbreak")
```

```{r}
library(Seurat)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(harmony)
library(purrr)
library(cowplot)
library(rstatix)
library(ggbeeswarm)
library(ggbreak)
```


# Functions
```{r}
# Define a function to run UMAP
create_UMAP <- function(seurat_obj, sample_name, num_pcs, res = 0.25, reduction = "pca", umap_dir = "~/figures/fig3/", processed_data_dir = "~/processed_data/") {

  # Find neighbors and clusters
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs)
  seurat_obj <- FindClusters(seurat_obj, resolution = res)
  
  # Print cluster assignments for verification
  print(paste("Clusters for", sample_name, ":"))
  print(head(Idents(seurat_obj)))
  
  # Run UMAP
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs, reduction = reduction)

  # Print message indicating completion
  print(paste("UMAP and clustering completed for:", sample_name))

  # Return the updated Seurat object
  return(seurat_obj)
}
```


```{r}

plot_marker_umaps <- function(seurat_obj, marker_csv, samplename = "sample", path_out = "~/figures/fig3/") {
  # Load marker gene list
  marker_df <- read_csv(marker_csv)
  
  # Validate columns
  if (!all(c("cell_type", "marker") %in% colnames(marker_df))) {
    stop("CSV must contain 'cell_type' and 'marker' columns")
  }
  
  # Drop NA or blank marker entries
  marker_df <- marker_df[!is.na(marker_df$marker) & marker_df$marker != "", ]
  
  plot_list <- list()

  # First: Add UMAP with Seurat cluster identity
  cluster_plot <- DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5) +
    ggtitle("Clusters")
  plot_list[[1]] <- cluster_plot

  # Then: Add FeaturePlots for each marker
  for (i in seq_len(nrow(marker_df))) {
    marker <- marker_df$marker[i]
    cell_type <- marker_df$cell_type[i]
    
    if (marker %in% rownames(seurat_obj)) {
      p <- FeaturePlot(seurat_obj, features = marker, reduction = "umap", pt.size = 0.25, order = TRUE) +
        ggtitle(paste0(cell_type, ": ", marker))
    } else {
      p <- ggplot() + 
        theme_void() + 
        ggtitle(paste0(cell_type, ": ", marker, " (not found)"))
    }
    plot_list[[i + 1]] <- p  # shift index by 1 to account for cluster plot
  }
  
  # Layout: adjust ncol to control grid width
  ncol <- 6
  nrow <- ceiling(length(plot_list) / ncol)
  final_plot <- wrap_plots(plotlist = plot_list, ncol = ncol)
  
  # Save
  #jpg_filename <- paste0(path_out, samplename, ".jpg")
  pdf_filename <- paste0(path_out, samplename, ".pdf", device = cairo_pdf())
  ggsave(pdf_filename, final_plot, width = 4 * ncol, height = 3 * nrow)
  
  print(final_plot)
  
  message("Saved plot to: ", pdf_filename)
}
```


# import and format seurat object
```{r}
srt_merge <- readRDS("~/processed_data/tumors_normalized_merged_scaled_dimreduction_annotated_srt_submission.rds")
```


```{r}
# add useful metadata
sample_name <- gsub("_", " ", srt_merge$orig.ident)
srt_merge <- AddMetaData(srt_merge, sample_name, col.name = "sample_name")

patient <- srt_merge@meta.data$sample_name %>% word()
srt_merge <- AddMetaData(srt_merge, patient, col.name = "patient")

Timepoint <- ifelse(grepl("Pre", srt_merge$sample_name), "Pre", "Post")
srt_merge <- AddMetaData(srt_merge, Timepoint, col.name = "Timepoint")
```

```{r}
# check metadata
srt_merge$sample_name %>% table()
srt_merge$patient %>% table()
srt_merge$Timepoint  %>% table()
```


# create custom cell colors
```{r}
# get cell types
cell_types <- unique(srt_merge@meta.data$cell_type_manual_refined)
# and order
desired_order_celltypes <- c("Malignant", "Neuron", "Glial cell", "Plasma cell", "CD4 T cell", "CD8 T cell", "Regulatory T cell", "Cycling lymphoid cell", "NK cell", "Macrophage", "Monocyte", "pDC", "Cycling myeloid cell", "Endothelial", "Mixed lymphoid-myeloid")

# make sure cell types in order vector and srt obj match
all(cell_types %in% desired_order_celltypes)
cell_types[!cell_types %in% desired_order_celltypes]
```


# set up colors
```{r}
named_colors_tumor <- c(
  "Malignant"              = "#B22222", # Tumor-exclusive cell types
  "CD4 T cell"             = "#A6CEE3", # CSF-overlapping cell types
  "CD8 T cell"             = "#B294C7", # CSF-overlapping cell types
  "Cycling T cell"         = "#C7B699", # CSF-overlapping cell types
  "Regulatory T cell"      = "#52AF43", # CSF-overlapping cell types
  "NK cell"                = "#E6CB75", # CSF-overlapping cell types
  "Macrophage"             = "#4992C2", # CSF-overlapping cell types
  "Monocyte"               = "#F68181", # CSF-overlapping cell types
  "pDC"                    = "#70449D", # CSF-overlapping cell types
  "Cycling myeloid cell"   = "#E6AB02", # Tumor-exclusive cell types
  "Plasma cell"            = "#AADB84", # CSF-overlapping cell types
  "Mixed lymphoid-myeloid" = "#4DAF4A", # Tumor-exclusive cell types
  "Glial cell"             = "#984EA3", # Tumor-exclusive cell types
  "Neuron"                 = "#377EB8", # Tumor-exclusive cell types
  "Endothelial"            = "#FF69B4" # Tumor-exclusive cell types
)

```


```{r}
# Set seed so UMAP runs the same every time. 
set.seed(1234)
```


```{r}
# set up batches:
batches <- case_when(srt_merge$sample_name == "P7 Post" ~ "Library batch 1",
                     srt_merge$sample_name == "P7 Pre" ~ "Library batch 2",
                     srt_merge$sample_name %in% c("P4 Pre", "P4 Post") ~ "Library batch 3",
                     srt_merge$sample_name %in% c("P1 Pre", "P1 Post") ~ "Library batch 4")
 
srt_merge <- AddMetaData(srt_merge, batches, col.name = "batch")
```



```{r}
# get cell counts for types and make new metadata column
# Count number of cells per group
cell_counts <- table(srt_merge$cell_type_manual_refined)

# Create a named vector of labels with counts
label_with_counts <- paste0(names(cell_counts), " (", cell_counts, ")")
names(label_with_counts) <- names(cell_counts)

# Add a new metadata column with those labels
counted_cells <- label_with_counts[match(srt_merge$cell_type_manual_refined, names(label_with_counts))]
names(counted_cells) <- colnames(srt_merge)
srt_merge <- AddMetaData(srt_merge, metadata = counted_cells, col.name = "cell_type_with_counts")

# check cell types with counts
srt_merge$cell_type_with_counts %>% table()
```


```{r}
# set up colors
named_colors_tumor <- c(
  "Malignant (27312)"              = "#B22222",
  "CD4 T cell (1808)"              = "#A6CEE3",
  "CD8 T cell (4181)"              = "#B294C7",
  "Cycling lymphoid cell (69)"     = "#C7B699",
  "Regulatory T cell (176)"        = "#52AF43",
  "NK cell (764)"                  = "#E6CB75",
  "Macrophage (12506)"             = "#4992C2",
  "Monocyte (3230)"                = "#F68181",
  "pDC (106)"                      = "#70449D",
  "Cycling myeloid cell (331)"     = "#E6AB02",
  "Plasma cell (68)"               = "#AADB84",
  "Mixed lymphoid-myeloid (273)"   = "#4DAF4A",
  "Glial cell (9827)"              = "#984EA3",
  "Neuron (77)"                    = "#377EB8",
  "Endothelial (295)"              = "#FF69B4"
)

```

```{r}
patient_colors <- c(
  "P1 Pre (17397)"  = "orange",  
  "P1 Post (2602)" =  "#e97132", 
  "P4 Pre (12375)"  = "lightgreen",  
  "P4 Post (1553)"  = "#196b24",  
  "P7 Pre (12920)"  = "#0f9ed5",  
  "P7 Post (14176)" = "#156082"   
)
```

```{r}
# get cell counts for samples and make new metadata column
# Count number of cells per group
cell_counts <- table(srt_merge$sample_name)
# Create a named vector of labels with counts
label_with_counts <- paste0(names(cell_counts), " (", cell_counts, ")")
names(label_with_counts) <- names(cell_counts)
label_with_counts
# Add a new metadata column with those labels
counted_cells <- label_with_counts[match(srt_merge$sample_name, names(label_with_counts))]
names(counted_cells) <- colnames(srt_merge)

srt_merge <- AddMetaData(srt_merge, metadata = counted_cells, col.name = "sample_name_with_counts")

srt_merge$sample_name_with_counts %>% table()
```




```{r}
# factor for order
srt_merge$cell_type_with_counts <- factor(srt_merge$cell_type_with_counts,
                                             levels = names(named_colors_tumor))

srt_merge$sample_name_with_counts <- factor(srt_merge$sample_name_with_counts, 
                                            levels = names(patient_colors))
```

# Fig 3B,C - all tumor cell UMAP
Fig. 3B: UMAP of single-nucleus transcriptomes from matched pre- and post-treatment GBM tissues across three patients.
```{r}
plot1 <- DimPlot(srt_merge, group.by = "seurat_clusters", label = TRUE, raster = FALSE) + 
  ggtitle("Paired tumors pre and post CAR T cell infusion")

plot2 <- DimPlot(srt_merge, group.by = "cell_type_manual_refined", raster = FALSE) + 
  ggtitle("Paired tumors pre and post CAR T cell infusion by cell type")

plot3 <- DimPlot(srt_merge, group.by = "sample_name_with_counts", label = TRUE, raster = FALSE,
                 cols = patient_colors) + 
  ggtitle("Paired tumors pre and post CAR T cell infusion")

plot4 <- DimPlot(srt_merge, group.by = "Timepoint", label = FALSE) + 
  ggtitle("Paired tumors pre and post CAR T cell infusion by infusion")

plot5 <- DimPlot(srt_merge, group.by = "cell_type_with_counts", 
                  cols = named_colors_tumor, raster = FALSE) +
  ggtitle("Paired tumors pre and post CAR T cell infusion by cell type")

#ggsave(filename = "~/figures/fig3/Fig3C_Tumors_merged_UMAP_colored_by_sample.pdf", plot = plot3, width = 14, height = 10)
#ggsave(filename = "~/figures/fig3/Fig3B_Tumors_merged_UMAP_colored_by_celltype_unifiedcols.pdf", plot = plot5, width = 14, height = 10)

plot1
plot2
plot3
plot4
plot5
```

```{r}
# establish sample order again. important for next plot as lacks cell count
patient_colors <- c(
  "P1 Pre"  = "orange",  
  "P1 Post" = "#e97132", 
  "P4 Pre"  = "lightgreen",  
  "P4 Post" = "#196b24",  
  "P7 Pre"  = "#0f9ed5",  
  "P7 Post" = "#156082"   
)
```


#Supp Fig 4F: plot sample-wise cell type composition
```{r}
# Extract metadata
meta <- srt_merge@meta.data

# Create a summary table with proportions
composition_df <- meta %>%
  group_by(sample_name, cell_type_with_counts) %>% # make a separate dataframe for each unique combination of values of orig.ident and cell_type_manual and sends to...
  summarise(n = n()) %>% # this counts up the number of entries (cells) for each combination and outputs them in a new column "n"
  ungroup() %>%
  group_by(sample_name) %>% # for each value of sample_name
  mutate(proportion = n/sum(n)) # for each value of cell type, give (count of that cell type)/(count of all cell types)

composition_df$cell_type_with_counts <- factor(composition_df$cell_type_with_counts, levels = names(named_colors_tumor))
composition_df$sample_name <- factor(composition_df$sample_name, levels = names(patient_colors))

# Plot
p <- ggplot(composition_df, aes(x = sample_name, y = proportion, fill = cell_type_with_counts)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = named_colors_tumor) +
  ylab("Proportion of Cells") +
  xlab("Sample") +
  labs(fill = "Cell Type") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) 

#ggsave(plot = p, filename = "~/figures/fig3/SuppFig3X_tumor_cell_types_stacked_barplot.pdf", width = 5, height = 5)
p
```

# Fig 3E, Supp Fig 4G plot T cell and myeloid  types by timepoint
```{r}
# set up colors
tcell_colors <- c(
  "CD4 T cell"              = "#A6CEE3",
  "CD8 T cell"              = "#B294C7",
  "Regulatory T cell"        = "#52AF43"
)

```

```{r}
# set up colors
myeloid_colors <- c(
  "Macrophage"           = "#4992C2",
  "Monocyte"             = "#F68181",
  "pDC"                  = "#70449D"
)

```

```{r}
immune_cells <- c(
  "CD4 T cell", "CD8 T cell", "Cycling lymphoid cell", "Regulatory T cell", "NK cell", "Macrophage", "Monocyte","pDC", "Cycling myeloid cell", "Plasma cell", "Mixed lymphoid-myeloid"
)
```

```{r}
# get composition of immune cell types
composition_df <- meta %>%
  filter(cell_type_manual_refined %in% immune_cells) %>%
  group_by(Timepoint, cell_type_manual_refined) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Timepoint) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

# get t cell fractions
tcell_df <- composition_df %>%
  filter(cell_type_manual_refined %in% names(tcell_colors))

# Compute "Other" as 1 - sum of T cells for each timepoint
other_df <- tcell_df %>%
  group_by(Timepoint) %>%
  summarize(proportion = 1 - sum(proportion)) %>%
  mutate(cell_type_manual_refined = "Other immune cells")

# Combine back with T cell data
plot_df_tcell <- bind_rows(tcell_df, other_df)

# Add gray color for "Other immune cells"
tcell_colors_full <- c(tcell_colors, "Other immune cells" = "gray70")

# Ensure factors are in the correct order
plot_df_tcell$cell_type_manual_refined <- factor(plot_df_tcell$cell_type_manual_refined, levels = c( "Other immune cells", names(tcell_colors)))
plot_df_tcell$Timepoint <- factor(plot_df_tcell$Timepoint, levels = c("Pre", "Post"))

# get myeloid fractions
myeloid_df <- composition_df %>%
  filter(cell_type_manual_refined %in% names(myeloid_colors))

# Compute "Other" as 1 - sum of T cells for each timepoint
other_df <- myeloid_df %>%
  group_by(Timepoint) %>%
  summarize(proportion = 1 - sum(proportion)) %>%
  mutate(cell_type_manual_refined = "Other immune cells")

# Combine back with T cell data
plot_df_myeloid <- bind_rows(myeloid_df, other_df)

# Add gray color for "Other immune cells"
myeloid_colors_full <- c(myeloid_colors, "Other immune cells" = "gray70")

# Ensure factors are in the correct order
plot_df_myeloid$cell_type_manual_refined <- factor(plot_df_myeloid$cell_type_manual_refined, levels = c( "Other immune cells", names(myeloid_colors)))
plot_df_myeloid$Timepoint <- factor(plot_df_myeloid$Timepoint, levels = c("Pre", "Post"))
```

```{r}
p1 <- ggplot(plot_df_tcell, aes(x = Timepoint, y = proportion, fill = cell_type_manual_refined)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = tcell_colors_full) +
  ylab("Proportion of Immune Cells") +
  xlab("Timepoint") +
  labs(fill = "Cell Type",
       title = "Proportion of T cells (out of immune cells") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_break(c(0.3, 0.9))

p2 <- ggplot(plot_df_myeloid, aes(x = Timepoint, y = proportion, fill = cell_type_manual_refined)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = myeloid_colors_full) +
  ylab("Proportion of Immune Cells") +
  xlab("Timepoint") +
  labs(fill = "Cell Type",
       title = "Proportion of Myeloid cells (out of immune cells") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) 


#ggsave(plot = p1, filename = "~/figures/fig3/Fig3E_Tcell_stacked_barplot_with_all_immune.pdf", width = 5, height = 7)
#ggsave(plot = p2, filename = "~/figures/suppfig4/SuppFig4G_myeloid_stacked_barplot_with_all_immune.pdf", width = 5, height = 7)

p1
p2
```


# Supp Table 3 - tumor sample characteristics
```{r}
srt_merge$cell_type_manual_refined %>% unique() 
immune_cell_types <- c("Macrophage", "Monocyte", "CD8 T cell", "CD4 T cell", "Regulatory T cell", "NK cell", "pDC",  "Cycling myeloid cell", "Cycling lymphoid cell", "Plasma cell", "Mixed lymphoid-myeloid")

supptab3 <- srt_merge@meta.data %>%
  group_by(sample_name) %>%
  summarize(total_cell_count = n(),
            nMalig = sum(cell_type_manual_refined == "Malignant"),
            nImmune = sum(cell_type_manual_refined %in% immune_cell_types),
            malig_immune_ratio = nMalig/nImmune
            )
supptab3
#write_tsv(supptab3, "~/figures/supp_table_3.tsv")
```

# Supp Fig 4B -Tumors colored by batch
```{r}
# suppose we want separate plots for each batch
umap_df <- as.data.frame(Embeddings(srt_merge, "umap")) %>%
  rownames_to_column("cell_id") %>%
  left_join(srt_merge@meta.data %>% rownames_to_column("cell_id"), by = "cell_id")

cols <- c("Library batch 1" = "#1F78B4", "Library batch 2" = "#FF7F00", "Library batch 3" = "#33A02C",  "Library batch 4" = "#E31A1C")

batch_levels <- names(cols)

plots <- map(batch_levels, function(b) {
  ggplot() +
    geom_point(data = umap_df, aes(x = umap_1, y = umap_2), color = "gray90", size = 0.5) +
    geom_point(data = filter(umap_df, batch == b), 
               aes(x = umap_1, y = umap_2, color = batch), size = 0.6) +
    scale_color_manual(values = cols) +
    ggtitle(paste(b)) +
    theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "none",
      panel.grid = element_blank()
    )
})

p <- plot_grid(plotlist = plots, ncol = 2)

#ggsave("~/figures/fig3/SuppFig4B_TumorCells_colored_by_LP_batch_faceted.pdf", plot = p, width = 5, height = 5)
p
```


# Supp Fig 4D - DotPlot Cell type markers by cell type
Cell type annotations based on canonical lineage marker expression.

```{r}
# clean up dotplot data to remove unwanted cell types
srt_merge_dotplot <- srt_merge
srt_merge_dotplot <- subset(srt_merge_dotplot, subset = cell_type_manual_refined != "Malignant")

# combine cycling cells
srt_merge_dotplot$cell_type_manual_refined[(srt_merge_dotplot$cell_type_manual_refined == "Cycling myeloid cell" 
                                            | srt_merge_dotplot$cell_type_manual_refined == "Cycling lymphoid cell" )] <- "Cycling cell"

srt_merge_dotplot$cell_type_manual_refined %>% table()
```

```{r, fig.width = 10, fig.height = 4, dpi = 180}
# output directory
out_dir <- "~/figures/fig3/"

# Read in the marker list once
markers_df <- read_csv("~/resources/cell_type_markers_forDotplot_refined.csv")
marker_genes <- unique(markers_df$marker)
marker_genes <- factor(marker_genes, levels = marker_genes)


# Define the desired cell type order
desired_order <- c("Neuron", "Glial cell", "CD8 T cell", "CD4 T cell", "Regulatory T cell", "NK cell", "Plasma cell", "Mixed lymphoid-myeloid", "Monocyte", "Macrophage", "pDC", "Endothelial", "Cycling cell")

# check that the ordered cell types are identical to the values in metadata
all(desired_order %in% unique(srt_merge_dotplot@meta.data$cell_type_manual_refined))
all(unique(srt_merge_dotplot@meta.data$cell_type_manual_refined) %in% desired_order)

srt_cells <- unique(srt_merge_dotplot$cell_type_manual_refined)
srt_cells[!srt_cells %in% desired_order]

srt_merge_dotplot@meta.data$cell_type_manual_refined <- factor(srt_merge_dotplot@meta.data$cell_type_manual_refined, levels = desired_order)

  
plot <- DotPlot(srt_merge_dotplot,  features = marker_genes, group.by= 'cell_type_manual_refined', dot.scale = 6, cluster.idents = FALSE) + RotatedAxis() +
  scale_color_viridis_c(option = "D", name = "Mean scaled expression") 

#ggsave("~/figures/suppfig4/SuppFig4B_DotPlot_AllCells_RefinedMarkers.pdf", plot = plot, width = 12, height = 6, device = cairo_pdf())
plot
```


```{r}
#remove dotplot to save space
rm(srt_merge_dotplot)
```

# Supp Fig 4C: UMAP - colored by marker genes
```{r}
markers_path <- "~/resources/cell_type_markers_for_multi_marker_UMAP.csv"

#plot_marker_umaps(srt_merge, markers_path, samplename = "SuppFig4C_UMAP_colored_by_marker_genes", path_out = "~/figures/suppfig4/")
plot_marker_umaps(srt_merge, markers_path, samplename = "SuppFig4C_UMAP_colored_by_marker_genes", path_out = "~/figures/")
```


# Supp Fig 4J - Malignant cells: EGFR UMAP
EGFR expression in neoplastic cells pre- and post-CAR T treatment – all patients together? On the UMAP?.

```{r}
srt_merge_malig <- subset(srt_merge, subset = cell_type_manual_refined == "Malignant")
```

```{r}
# get cell counts for timepoint and make new metadata column

# Count number of cells per group
cell_counts <- table(srt_merge_malig$Timepoint)

# Create a named vector of labels with counts
label_with_counts <- paste0(names(cell_counts), " (", cell_counts, ")")
names(label_with_counts) <- names(cell_counts)

# Add a new metadata column with those labels
counted_cells <- label_with_counts[match(srt_merge_malig$Timepoint, names(label_with_counts))]
names(counted_cells) <- colnames(srt_merge_malig)

srt_merge_malig <- AddMetaData(srt_merge_malig, metadata = counted_cells, col.name = "timepoint_with_counts")
srt_merge_malig$timepoint_with_counts %>% table()
```

```{r}
tp_cols <- c(
  "Pre (15632)" = "#A6BD8C", 
  "Post (11680)" = "firebrick4")

srt_merge_malig$timepoint_with_counts <- factor(srt_merge_malig$timepoint_with_counts, 
                                                levels = names(tp_cols))
srt_merge_malig$timepoint_with_counts %>% table()
```


```{r}
p1 <- FeaturePlot(srt_merge_malig, features = "EGFR", raster = FALSE) + 
  ggtitle("Malignant Cells: EGFR expression")
p2 <- DimPlot(srt_merge_malig, group.by = "timepoint_with_counts", cols = tp_cols, raster = FALSE) + 
  ggtitle("Malignant Cells: Treatment")
p3 <- DimPlot(srt_merge_malig, group.by = "sample_name") + 
  ggtitle("Malignant Cells: Sample")

#ggsave(filename = "~/figures/suppfig4/SuppFig4J_MaligCells_EGFR_UMAP_EGFR_expression.pdf", plot = p1, width = 8, height = 6)
#ggsave(filename = "~/figures/suppfig4/SuppFig4J_MaligCells_EGFR_UMAP_by_timepoint.pdf", plot = p2, width = 8, height = 6)
#ggsave(filename = "~/figures/suppfig4/MaligCells_UMAP_by_sample.pdf", plot = p3, width = 8, height = 6)

p1
p2
p3
```


# Supplemental Figure 3G - EGFR expression by sample
```{r}

# setup EGFR boxplots
# Get EGFR and IL13RA2 expression and metadata
malig_df <- FetchData(srt_merge_malig, vars = c("EGFR", "IL13RA2", "sample_name", "Timepoint"))
colnames(malig_df)[1] <- "EGFR"  # make sure it's labeled clearly

comparisons <- list(
  c("P1 Pre", "P1 Post"),
  c("P4 Pre", "P4 Post"),
  c("P7 Pre", "P7 Post")
)
unique(malig_df$sample_name)
```

```{r}
patient_colors <- c(
  "P1 Pre"  = "orange",  
  "P1 Post" = "#e97132", 
  "P4 Pre"  = "lightgreen",  
  "P4 Post" = "#196b24",  
  "P7 Pre"  = "#0f9ed5",  
  "P7 Post" = "#156082"   
)
```

```{r}
sample_order <- c("P1 Pre", "P1 Post", "P7 Pre", "P7 Post", "P4 Pre", "P4 Post")
malig_df$sample_name <- factor(malig_df$sample_name, levels = sample_order)
malig_df$sample_name %>% table()

all(names(patient_colors) %in% malig_df$sample_name)
```
# Fig 3H - plot of egfr levels by sample (Pt/Tp) 

```{r}
# violin plot + beeswarm
plot <- ggplot(malig_df, aes(x = sample_name, y = EGFR, fill = sample_name)) +
  geom_violin(trim = FALSE, scale = "width", color = NA, alpha = 0.8) +  # violin plot
  geom_beeswarm(priority = "density", cex = 0.1, alpha = 0.2, color = "gray") +  # beeswarm overlay
  scale_fill_manual(values = patient_colors) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4, color = "black", flatten = 0) +  # mean point
  stat_compare_means(comparisons = comparisons, method = "wilcox.test", label = "p.format") +
  theme_minimal() +
  labs(title = "Malignant cells: EGFR expression by sample", x = "Sample", y = "EGFR Expression") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.position = "none")

#ggsave("~/figures/fig3/Fig3F_MaligCell_EGFR_expression_by_sample_wilcox_violin_beeswarm.pdf", plot = plot)
plot
```



# plot interval between infusion and second surgery
```{r}

timeline_df <- read_csv("~/external_resources/surgery_interval.csv")

timeline_df$Patient <- factor(timeline_df$Patient, levels = c("P1", "P7", "P4"))
colors = c("P1" = "#e97132", "P4" = "#196b24", "P7" = "#0f9ed5")

plot <- ggplot(timeline_df, aes(x = Patient, y = infusion_second_surgery_interval, fill = Patient)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none")

#ggsave("~/figures/fig3/Fig3G_patient_infusion_surgery_interval.pdf", width = 2.5, height = 5, plot = plot)

plot
```

# Fig 3I Plot PFS
```{r}
colors = c("P1" = "#e97132", "P4" = "#196b24", "P7" = "#0f9ed5")
pfs <- read_csv("~/external_resources/PFSforDana_latest_.csv") %>%
  dplyr::filter(patient %in% c("P1", "P7", "P4")) 
  
pfs$patient <- factor(pfs$patient, levels = c("P1", "P7", "P4"))

plot <- ggplot(pfs, aes(x = patient, y = pfs, fill = patient)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  labs(y = "Progression-free survival",
       title = "Progression-free survival by patient") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none")
#ggsave("~/figures/fig3/Fig3G_pfs.pdf", width = 2.5, height = 5, plot = plot)  
plot
```

# Fig 3I -plot CSF CD8 CAR T cell cytotoxicity and exhaustion score by sample 
```{r}
sample_colors = c("P1D7" = "#e97132", "P4D7" = "#196b24", "P7D7" = "#0f9ed5")
srt_csf_CAR_CD8 <- readRDS("~/external_resources/CSF_scData/annotated_prepost_csf.rds") %>%
  subset(subset = cell_type == "CD8+ CAR T") %>%
  subset(subset = sample_id %in% names(sample_colors))

# read in and format signature
sigs <- read_csv("~/external_resources/wherry_pace_signatures.csv")
markers <- list("Pace effector markers" = sigs$Pace_effector_markers[!is.na(sigs$Pace_effector_markers)],
                              "Wherry exhaustion markers" = sigs$Wherry_exhaustion_markers[!is.na(sigs$Wherry_exhaustion_markers)])

srt_csf_CAR_CD8 <- AddModuleScore(srt_csf_CAR_CD8,
                                features = markers)

srt_csf_CAR_CD8$sample_id <- factor(srt_csf_CAR_CD8$sample_id, levels = names(sample_colors))

meta <- srt_csf_CAR_CD8@meta.data 
colnames(meta)[colnames(meta) == "Cluster1"] = "Pace effector markers"
colnames(meta)[colnames(meta) == "Cluster2"] = "Wherry exhaustion markers"
```


```{r}
# plot effector markers - D7 only
samples <- c("P1D7", "P7D7", "P4D7")
#sample_colors = c("P1D7" = "#e97132", "P4D7" = "#196b24", "P7D7" = "#0f9ed5")
meta <- meta %>% dplyr::filter(sample_id %in% samples)
meta$sample_id <- factor(meta$sample_id, levels = samples)

comparisons_list <- list(c("P4D7", "P1D7"), 
                         c("P4D7", "P7D7"), 
                         c("P1D7", "P7D7"))

plot1 <- ggplot(data = meta, aes(x = sample_id, y = `Pace effector markers`, fill = sample_id)) +
  geom_violin() + 
  scale_fill_manual(values = sample_colors) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.75, color = "black", fatten = 1) + 
  stat_compare_means(comparisons = comparisons_list, 
                     method = "wilcox.test",
                     label = "p.format", 
                     hide.ns = TRUE) +
  theme_bw() +
  labs(title = "CD8 CAR T-D7",
       subtitle = "Pace Effector",
       x = "Pace Effector Markers expression") +
  theme(panel.grid = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none")

plot2 <- ggplot(data = meta, aes(x = sample_id, y = `Wherry exhaustion markers`, fill = sample_id)) +
  geom_violin() + 
  scale_fill_manual(values = sample_colors) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.75, color = "black", fatten = 1) +  
  stat_compare_means(comparisons = comparisons_list, 
                     method = "wilcox.test",
                     label = "p.format", 
                     hide.ns = TRUE) +
  theme_bw() +
  labs(title = "CD8 CAR T-D7",
       subtitle = "Wherry Exhaustion",
       x = "Wherry exhaustion marker expression") +
  theme(panel.grid = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none")


#ggsave("~/figures/fig3/Fig3G_CD8_CAR_Tcells_D7_Pace_effector_markers.pdf", plot = plot1, width = 3, height = 6)
#ggsave("~/figures/fig3/Fig3G_CD8_CAR_Tcells_D7_Wherry_exhaustion_markers.pdf", plot = plot2, width = 3, height = 6)
plot1
plot2
```

```{r}
# externally calculate wilcox tests for effector/cytotoxicity and exhaustion markers
# Define comparisons
samples <- c("P1D7", "P4D7", "P7D7")
comparisons <- combn(samples, 2, simplify = FALSE)  # all pairwise

# Function to do Wilcoxon tests for a given signature column
run_wilcox_tests <- function(df, column) {
  df %>%
    dplyr::select(sample_id, !!sym(column)) %>%
    wilcox_test(as.formula(paste0("`", column, "` ~ sample_id"))) %>%
    adjust_pvalue(method = "bonferroni") %>%
    add_significance()
}

# For Pace effector markers
wilcox_effector <- run_wilcox_tests(meta, "Pace effector markers")

# For Wherry exhaustion markers
wilcox_exhaustion <- run_wilcox_tests(meta, "Wherry exhaustion markers")
```


# CSF Tumor integration
```{r}
# Import CSF
external_path <- "~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis"
srt_csf <- readRDS(file.path(external_path, "external_resources/CSF_scData/annotated_prepost_csf.rds"))


# integrate cell types
srt_csf$cell_type_manual_refined <- srt_csf$cell_type

srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "CD4+ T"] <- "CD4 T cell"
srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "CD8+ T"] <- "CD8 T cell"
srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "Plasma"] <- "Plasma cell"
srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "Cycling CD8+ T"] <- "Cycling lymphoid cell"
srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "Cycling CD4+ T"] <- "Cycling lymphoid cell"
srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "Treg"] <- "Regulatory T cell"
srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "B"] <- "B cell"
srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "NK"] <- "NK cell"
srt_csf$cell_type_manual_refined[srt_csf$cell_type_manual_refined == "Cycling T cell"] <- "Cycling lymphoid cell"

srt_csf$cell_type_manual_refined %>% table()
```


```{r}
# import tumors
srt_merge <- readRDS("~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/processed_data/tumors_normalized_merged_scaled_dimreduction_annotated_srt_submission.rds")
#srt_merge <- readRDS("~/processed_data/tumors_normalized_merged_scaled_dimreduction_annotated_withTCR_srt.rds")

srt_merge$cell_type_manual_refined %>% table()
```


```{r}
srt_tumor_csf <- merge(srt_merge, y = srt_csf)

# In Seurat v5 you have to manually join all the layers.
# separately join normalized layers (data) and raw counts "counts"

srt_tumor_csf <- JoinLayers(srt_tumor_csf)
Layers(srt_tumor_csf)
```


```{r}
# Run UMAP on combined data
set.seed(1234)
sample_name <- "tumors_csf_combined"

out_dir <- file.path(external_path, "results/dimreduction/umap")

# find variable features
srt_tumor_csf <- FindVariableFeatures(srt_tumor_csf, selection.method = "vst", nfeatures = 2000)

srt_tumor_csf <- ScaleData(srt_tumor_csf, features = VariableFeatures(srt_tumor_csf))

srt_tumor_csf <- RunPCA(srt_tumor_csf, features = VariableFeatures(object = srt_tumor_csf))

ElbowPlot(srt_tumor_csf)
```

From elbow plot, use 10 PCs
```{r}
srt_tumor_csf <- create_UMAP(srt_tumor_csf, sample_name = sample_name, umap_dir = out_dir, num_pcs = 10, res = 0.25)

DimPlot(srt_tumor_csf, group.by = "cell_type_manual_refined", raster = FALSE) +
  ggtitle("UMAP: CSF and Tumor cells - no batch correction")

#ggsave("~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/figures/fig3/Fig3D_UMAP_tumors_csf_combined.pdf", 
#       width = 14, height = 9, dpi = 300, units = "in")

srt_tumor_csf@meta.data %>% head()
srt_tumor_csf@meta.data$cell_type_manual_refined %>% table() 
srt_tumor_csf@meta.data$orig.ident %>% table()
```

# Prep data to examine batches of CSF - Tumor
```{r}
tissue_type <- case_when(srt_tumor_csf@meta.data$orig.ident == "SeuratProject" ~ "CSF",
                         TRUE ~ "Tumor")

srt_tumor_csf <- AddMetaData(srt_tumor_csf, tissue_type, col.name = "tissue_type")

```



```{r}
# batch correct with harmony
srt_tumor_csf <- RunHarmony(srt_tumor_csf, group.by.vars = "tissue_type")
```

```{r}
set.seed(1234)
sample_name <- "tumor_csf_harmony"
# run UMAP on harmony reduction
ElbowPlot(srt_tumor_csf, reduction = "harmony")
# from elbowplot, use 11 dims

srt_tumor_csf <- create_UMAP(srt_tumor_csf, sample_name = sample_name, reduction = "harmony", umap_dir = out_dir, num_pcs = 11, res = 0.25)

```

```{r}
# get cell counts for types and make new metadata column
# Count number of cells per group
cell_counts <- table(srt_tumor_csf$cell_type_manual_refined)
cell_counts
# Create a named vector of labels with counts
label_with_counts <- paste0(names(cell_counts), " (", cell_counts, ")")
names(label_with_counts) <- names(cell_counts)
label_with_counts
# Add a new metadata column with those labels
counted_cells <- label_with_counts[match(srt_tumor_csf$cell_type_manual_refined, names(label_with_counts))]
names(counted_cells) <- colnames(srt_tumor_csf)
srt_tumor_csf <- AddMetaData(srt_tumor_csf, metadata = counted_cells, col.name = "cell_type_with_counts")

```


```{r}
# get cell counts for tissue types and make new metadata column
# Count number of cells per group
cell_counts <- table(srt_tumor_csf$tissue_type)
# Create a named vector of labels with counts
label_with_counts <- paste0(names(cell_counts), " (", cell_counts, ")")

names(label_with_counts) <- names(cell_counts)

# Add a new metadata column with those labels
counted_cells <- label_with_counts[match(srt_tumor_csf$tissue_type, names(label_with_counts))]
names(counted_cells) <- colnames(srt_tumor_csf)
srt_tumor_csf <- AddMetaData(srt_tumor_csf, metadata = counted_cells, col.name = "tissue_type_with_counts")


srt_tumor_csf$tissue_type_with_counts[1:10]
table(srt_tumor_csf$tissue_type_with_counts)
```


```{r}
named_colors_tissue <- c(
  "Tumor (61023)" = "#ff7f0e",
  "CSF (87386)" = "#589ea2"
)
```

```{r}
table(srt_tumor_csf$cell_type_with_counts)
```


```{r}
#set up colors with counts
named_colors_all <- c(
  # CSF cell types
  "CD4+ CAR T (17798)"          = "#569EA4",
  "CD4 T cell (21033)"          = "#A6CEE3",
  "CD8+ CAR T (13197)"          = "#D46F84",
  "CD8 T cell (13366)"          = "#B294C7",
  "Cycling lymphoid cell (774)" = "#C7B699",
  "Mixed CAR+/CAR- Tfh (1394)"  = "#FE8B14",
  "Regulatory T cell (7804)"    = "#52AF43",
  "NK cell (4553)"              = "#E6CB75",
  "MAIT (87)"                   = "#B15928",
  "Monocyte (8690)"             = "#F68181",
  "Macrophage (15960)"          = "#4992C2",
  "cDC1 (393)"                  = "#FDB761",
  "cDC2 (1108)"                 = "#8BAD49",
  "mregDC (1244)"               = "#FD8C4C",
  "pDC (961)"                   = "#70449D",
  "B cell (1358)"               = "#E5292B",
  "Plasma cell (574)"           = "#AADB84",

  # Tumor-exclusive cell types (beautified)
  "Malignant (27312)"              = "#B22222",
  "Cycling myeloid cell (331)"     = "#E6AB02",
  "Glial cell (9827)"              = "#984EA3",
  "Mixed lymphoid-myeloid (273)"   = "gray",
  "Endothelial (295)"              = "#FF69B4",
  "Neuron (77)"                    = "#377EB8"
)

```

# checkpoint - save harmony tumor_csf obj and reload
```{r}
# warning - only do this if necessary - very large file
saveRDS(srt_tumor_csf, "~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/processed_data/tumors_csf_combined_full_annotation_harmony_srt_submission.rds")
```

 subset merged tumor/CSF object with even numbers of tumor m
```{r}
# complex subsetting operation to get even numbers of CSF myleoid and tumor myelid and CSF lymphoid and tumor lymphoid cells
#myeloid_celltypes <- c("Monocyte", "cDC1", "cDC2", "mregDC", "pDC", "Cycling myeloid cell")
myeloid_celltypes <- c("Macrophage","Monocyte", "cDC1", "cDC2", "mregDC", "pDC", "Cycling myeloid cell")
lymphoid_celltypes <- c("CD4+ CAR T", "CD4 T cell", "CD8+ CAR T", "CD8 T cell", "Cycling T cell", "Mixed CAR+/CAR- Tfh", "Regulatory T cell", "NK cell", "MAIT")
other_celltypes <- unique(srt_tumor_csf$cell_type_manual_refined)[!unique(srt_tumor_csf$cell_type_manual_refined) %in% c(myeloid_celltypes, lymphoid_celltypes)]

othercells <- srt_tumor_csf %>% 
  subset(cell_type_manual_refined %in% other_celltypes) %>%
  colnames()

tumor_myeloid_cells <- srt_tumor_csf %>% 
  subset(cell_type_manual_refined %in% myeloid_celltypes) %>%
  subset(tissue_type == "Tumor") %>%
  colnames()

CSF_myeloid_cells <- srt_tumor_csf %>% 
  subset(cell_type_manual_refined %in% myeloid_celltypes) %>%
  subset(tissue_type == "CSF") %>%
  colnames()

tumor_lymphoid_cells <- srt_tumor_csf %>% 
  subset(cell_type_manual_refined %in% lymphoid_celltypes) %>%
  subset(tissue_type == "Tumor") %>%
  colnames()

CSF_lymphoid_cells <- srt_tumor_csf %>% 
  subset(cell_type_manual_refined %in% lymphoid_celltypes) %>%
  subset(tissue_type == "CSF") %>%
  colnames()



CSF_myeloid_count <- length(CSF_myeloid_cells)
tumor_lymphoid_count <- length(tumor_lymphoid_cells)

length(tumor_myeloid_cells)
length(CSF_myeloid_cells)
length(tumor_lymphoid_cells)
length(CSF_lymphoid_cells)
```


```{r}
#tumor_myeloid_subset <- sample(tumor_myeloid_cells, length(tumor_myeloid_cells)/5) # limit to length of CSF myeloid cells
tumor_myeloid_subset <- sample(tumor_myeloid_cells, length(tumor_myeloid_cells)/3) # limit to length of CSF myeloid cells
CSF_myeloid_subset <- sample(CSF_myeloid_cells, length(CSF_myeloid_cells))

tumor_lymphoid_subset <- sample(tumor_lymphoid_cells, length(tumor_lymphoid_cells)) 
CSF_lymphoid_subset <- sample(CSF_lymphoid_cells, length(tumor_lymphoid_cells)) # limit length to length of tumor lymphoid cells

#tumor_macrophage_subset <- sample(tumor_macrophages, length(tumor_macrophages)/5)

other_cell_subset <- sample(othercells, length(othercells))
```

```{r}
srt_tumor_csf_sub <- subset(srt_tumor_csf, cells = c(tumor_myeloid_subset,
                                                     tumor_lymphoid_cells,
                                                     CSF_myeloid_subset,
                                                     CSF_lymphoid_subset,
                                                     #tumor_macrophage_subset,
                                                     #CSF_macrophages,
                                                     other_cell_subset))
srt_tumor_csf_sub$cell_type_with_counts <- factor(srt_tumor_csf_sub$cell_type_with_counts, levels = names(named_colors_all))
srt_tumor_csf_sub$tissue_type_with_counts <- factor(srt_tumor_csf_sub$tissue_type_with_counts, levels = names(named_colors_tissue))
gc()
```


# Fig3F downsampled UMAP colored by tissue type
```{r fig.width= 7, fig.height=4}
p1 <- DimPlot(srt_tumor_csf_sub, group.by = "tissue_type_with_counts", cols = named_colors_tissue, shuffle = TRUE, raster = FALSE, stroke.size = .33) 
#ggsave("~/figures/fig3/Fig3F_UMAP_tumor_CSF_merge_by_tissue_downsample.pdf", plot = p1, width = 8, height = 6)
p1

```
# Fig3G downsampled UMAP colored by cell type
```{r fig.width = 14, fig.height= 5}
p2 <- DimPlot(srt_tumor_csf_sub, group.by = "cell_type_with_counts", cols = named_colors_all, shuffle = TRUE, raster = FALSE, stroke.size = 0.33) 
#ggsave("~/figures/fig3/Fig3G_UMAP_tumor_CSF_merge_celltype_downsample.pdf", plot = p2, width = 12, height = 6)
p2
```


