---
title: "Fig4_NKcells"
output: html_document
date: "2025-04-22"
---

```{r}
library(Seurat)
library(tidyr)
library(dplyr)
library(tibble)
library(ggplot2)
library(ggnewscale)
library(gghalves)
library(ggpubr) # needs cmake
library(scales)
library(harmony)
library(patchwork)
library(stringr)
library(forcats)
```


# Define functions
```{r}
# Define a function to run UMAP
create_UMAP <- function(seurat_obj, sample_name, num_pcs, res = 0.25, reduction = "pca", umap_dir = "~/results/dimreduction/umap/", processed_data_dir = "~/processed_data/") {

  # Find neighbors and clusters
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs, reduction = reduction)
  seurat_obj <- FindClusters(seurat_obj, resolution = res)
  
  # Print cluster assignments for verification
  print(paste("Clusters for", sample_name, ":"))
  print(head(Idents(seurat_obj)))
  
  # Run UMAP
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs, reduction = reduction)

  # Save UMAP plots with different groupings
  p1 <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE)

  print(p1)

  # Print message indicating completion
  print(paste("UMAP and clustering completed and files saved for:", sample_name))

  # Return the updated Seurat object
  return(seurat_obj)
}
```

# read in tumor sc data
```{r}
srt_merge <- readRDS("~/processed_data/tumors_normalized_merged_scaled_dimreduction_annotated_srt.rds")

# fix some cell type names
srt_merge$cell_type_manual_refined <- gsub("cells", "cell", srt_merge$cell_type_manual_refined)
srt_merge$cell_type_manual_refined[srt_merge$cell_type_manual_refined == "Proliferating myeloid cells"] <- "Cycling myeloid cell"
srt_merge$cell_type_manual_refined[srt_merge$cell_type_manual_refined == "Proliferating T cell"] <- "Cycling lymphoid cell"
srt_merge$cell_type_manual_refined[srt_merge$cell_type_manual_refined == "Ambiguous"] <- "Mixed lymphoid-myeloid"

# look at cell types
types <- srt_merge$cell_type_manual_refined %>% table() %>% names()

# remove P16
srt_merge <- subset(srt_merge, subset = orig.ident != "P16_Post_11907")

# for formality, remove redo from orig ident and colnames
srt_merge$orig.ident <- gsub("_redo", "", srt_merge$orig.ident)
colnames(srt_merge) <- gsub("_redo", "", colnames(srt_merge))

# fix sample labels for aesthetics. create as new labels for safety
sample_name <- gsub("_[0-9]{5}$", "", srt_merge$orig.ident)
sample_name <- gsub("_", " ", sample_name)
srt_merge <- AddMetaData(srt_merge, sample_name, col.name = "sample_name")

patient <- srt_merge@meta.data$sample_name %>% word()
srt_merge <- AddMetaData(srt_merge, patient, col.name = "patient")

Timepoint <- ifelse(grepl("Pre", srt_merge$sample_name), "Pre", "Post")
srt_merge <- AddMetaData(srt_merge, Timepoint, col.name = "Timepoint")

srt_merge$sample_name %>% table()
srt_merge$Timepoint %>% table()
srt_merge$patient %>% table()
colnames(srt_merge)[1:5]
```

```{r}
custom_colors <- c(
  "P1 Pre"  = "orange",  
  "P1 Post" = "#e97132", 
  "P4 Pre"  = "lightgreen",  
  "P4 Post" = "#196b24",  
  "P7 Pre"  = "#0f9ed5",  
  "P7 Post" = "#156082"   
)
```


# Subset NK cells
```{r}
# subset to NK cells and rerun UMAP
srt_merge_NK <- subset(srt_merge, subset = cell_type_manual_refined == "NK cell")

srt_merge_NK <- FindVariableFeatures(srt_merge_NK, selection.method = "vst", nfeatures = 2000)
srt_merge_NK <- ScaleData(srt_merge_NK, features = VariableFeatures(srt_merge_NK))
srt_merge_NK <- RunPCA(srt_merge_NK, features = VariableFeatures(object = srt_merge_NK))

ElbowPlot(srt_merge_NK)

```

Plot umap
```{r}
set.seed(1234)
srt_merge_NK <- create_UMAP(srt_merge_NK, "merged_nk_cells", num_pcs = 9, res = 0.25)
```

# annotate nk cell subtypes
```{r}
NK_type <- srt_merge_NK@meta.data %>% 
  mutate(NK_type = case_when(seurat_clusters == 1 ~ "CD56-Dim/CD16+",
                             seurat_clusters %in% c(0, 2) ~ "CD56-Bright/CD16-",
                             seurat_clusters == 3 ~ "other")) %>%
  pull(NK_type)

srt_merge_NK <- AddMetaData(srt_merge_NK, NK_type, "NK_type") 

```

# CHECKPOINT - save and reload NK cell seurat obj
```{r}
#saveRDS(srt_merge_NK, "~/processed_data/fig4_tumor_NK_only_type_annotated.rds")
```

```{r}
srt_merge_NK <- readRDS("~/processed_data/fig4_tumor_NK_only_type_annotated.rds")
```


# find nk type markers
```{r}
umap_nk_markers <- FindAllMarkers(srt_merge_NK, group.by = "NK_type")

topmarkers <- umap_nk_markers %>% 
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 20, with_ties = FALSE) %>%
  ungroup()

#write_tsv(topmarkers, "~/figures/fig4/FindAllMarkers_nk_cell_subtypes.tsv")
```


# add module scores for nk cell types - extended sigs
```{r}
CD16pos_CD56dim <- c("FCGR3A", "FGFBP2", "PRDM1", "CX3CR1", "CCL3", "CCL4, KIR3DL2", "KIR2DL3")
CD16neg_CD56bri <- c("NCAM1", "XCL1", "XCL2", "GZMK", "KLRC1", "SELL", "GATA3")

features <- list("CD16pos_CD56dim" = CD16pos_CD56dim,
                 "CD16neg_CD56bri" = CD16neg_CD56bri)

srt_merge_NK <- AddModuleScore(srt_merge_NK, features = features)
colnames(srt_merge_NK@meta.data)[colnames(srt_merge_NK@meta.data) %in% c("Cluster1", "Cluster2")] <- c("CD16pos_CD56dim", "CD16neg_CD56bri")
srt_merge_NK@meta.data
```

# Supp Fig5 F - CD56DimCD16+/CD56BrightCD16- Signature score over umap
```{r}
cols = c("#3c78b5", "#fdfab8", "#be2448")
p1 <- FeaturePlot(srt_merge_NK, features = "CD16pos_CD56dim", raster = FALSE) + 
  scale_color_gradientn(colors = cols, name = "Mean scaled expression") +
  ggtitle(label = "CD56-Dim/CD16-Pos Signature")


p2 <- FeaturePlot(srt_merge_NK, features = "CD16neg_CD56bri", raster = FALSE) + 
  scale_color_gradientn(colors = cols, name = "Mean scaled expression") +
  ggtitle(label = "CD16-Neg/CD56-Bright Signature")

#ggsave(filename = "~/figures/fig4/SuppFig5X_FeaturePlotCD16pos_CD56dim_big_sig.pdf", plot = p1, width = 8, height = 5.5)
#ggsave(filename = "~/figures/fig4/SuppFig5X_FeaturePlotCD16neg_CD56bri_big_sig.pdf", plot = p2, width = 8, height = 5.5)

p1
p2
```

# dotplot for expresssion of sigs by nk type
```{r}
greenpurple <- c("#036300", "white", "#68228b")

#Dotplot for CD56-Bright/CD56-Dim
features <- c(CD16pos_CD56dim, CD16neg_CD56bri)
features <- factor(features, levels = features)

srt_merge_NK_sub <- subset(srt_merge_NK, subset = NK_type != "other")

plot <- DotPlot(srt_merge_NK_sub, features = features, group.by = "NK_type") +
  scale_color_gradientn(colors = greenpurple) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
#ggsave("~/figures/fig4/Fig4D_DotPlot_CD56DCD16P_CD56BCD16N_green_purple.pdf", plot = plot, width = 8, height = 3.5)
plot
```


```{r}
# get cell counts for timepoint and make new metadata column

# Count number of cells per group
cell_counts <- table(srt_merge_NK$Timepoint)

# Create a named vector of labels with counts
label_with_counts <- paste0(names(cell_counts), " (", cell_counts, ")")
names(label_with_counts) <- names(cell_counts)

# Add a new metadata column with those labels
counted_cells <- label_with_counts[match(srt_merge_NK$Timepoint, names(label_with_counts))]
names(counted_cells) <- colnames(srt_merge_NK)

srt_merge_NK <- AddMetaData(srt_merge_NK, metadata = counted_cells, col.name = "timepoint_with_counts")
srt_merge_NK$timepoint_with_counts %>% table()
```

```{r}
tp_cols <- c(
  "Pre (477)" = "#A6BD8C", 
  "Post (285)" = "firebrick4")

srt_merge_NK$timepoint_with_counts <- factor(srt_merge_NK$timepoint_with_counts, 
                                                levels = names(tp_cols))
```

# Fig 4J plot pre/post cells over umap
```{r}
p1 <- DimPlot(srt_merge_NK, group.by = "timepoint_with_counts", cols = tp_cols, raster = FALSE)
#ggsave(filename = "~/figures/fig4/Fig4I_UMAP_NKcells_by_timepoint.pdf", plot = p1, width = 8, height = 5.5)
p1
```


# calculate ratio of cytox to sig NK cells - Tumor
```{r}
NK_type_fraction_by_sample <- srt_merge_NK@meta.data %>%
  group_by(sample_name, NK_type) %>%
  summarize(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = NK_type, values_from = n, values_fill = 0) %>%
  mutate(ratio = `CD56-Dim/CD16+`/`CD56-Bright/CD16-`)

desired_order_samples <- c("P1 Pre", "P1 Post", "P4 Pre", "P4 Post", "P7 Pre", "P7 Post")
NK_type_fraction_by_sample$sample_name <- factor(NK_type_fraction_by_sample$sample_name, levels = desired_order_samples)

NK_type_fraction_by_sample$timepoint <- ifelse(grepl("Pre", NK_type_fraction_by_sample$sample_name), "Pre-infusion", "Post-infusion")
NK_type_fraction_by_sample$timepoint <- factor(NK_type_fraction_by_sample$timepoint, levels = c("Pre-infusion", "Post-infusion"))

NK_type_fraction_by_sample$patient <- word(NK_type_fraction_by_sample$sample_name, 1)
NK_type_fraction_by_sample$patient <- factor(NK_type_fraction_by_sample$patient, levels = c("P1", "P4", "P7"))
```



# Fig 4I: plot NK subtype ratio for tumor by timepoint with lines, dots, bar
```{r}
plot_ratio_by_timepoint <- function(df, title_text, order) {
  # Compute mean per timepoint
  mean_df <- df %>%
    group_by(timepoint) %>%
    summarise(mean_ratio = mean(ratio, na.rm = TRUE), .groups = "drop")
  
  ggplot(df, aes(x = timepoint, y = ratio)) +
    geom_col(data = mean_df, aes(x = timepoint, y = mean_ratio), 
             fill = "gray30", width = 0.6, alpha = 0.5) +
    
    geom_point(data = df, aes(x = timepoint, y = ratio, color = patient), 
               size = 2) +
    
    geom_line(data = df, aes(x = timepoint, y = ratio, group = patient, color = patient), 
              linewidth = 0.7, alpha = 0.6) +

    scale_color_manual(values = c("P1" = "#e97132", 
                                  "P4" = "#196b24", 
                                  "P7" = "#0f9ed5")) +
    
    labs(x = "Timepoint", y = "CD56Dim / CD56Bright Ratio", 
         title = title_text) +
    theme_bw() +
    theme(panel.grid = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1))
}

```


```{r}

p_Tumor <- plot_ratio_by_timepoint(NK_type_fraction_by_sample, "Tumor NK Subtype Ratio  Across Timepoints") +
  stat_compare_means(data = NK_type_fraction_by_sample, 
                       mapping = aes(x = timepoint, y = ratio),  
                       method = "t.test", comparisons = list(c("Pre-infusion", "Post-infusion")), 
                       label = "p.format", hide.ns = FALSE) 

# Save 
#ggsave("~/figures/fig4/Fig4I_Tumor_ratio_plot.pdf", plot = p_Tumor, width = 5, height = 8)

# Display
p_Tumor

```

