---
title: "Fig_S4D_inferCNV.rmd"
output: html_document
date: "2026-02-17"
---


```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(readr)
library(infercnv)
```


# Define functions
This outputs annotations files for cell ranger cell ids and seurat clusters and returns a gene expression matrix
```{r}
# Define function to process each Seurat object with user-defined clusters
process_seurat_annotations <- function(seurat_obj, sample_name,
                                       output_dir = "~/resources/") {
  
  # Ensure output directory exists
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Extract gene expression matrix
  GEX_matrix <- as.matrix(GetAssayData(seurat_obj, assay = "RNA", slot = "counts"))
  
  # Create annotations for Cell Ranger and Seurat clusters
  annotations_seurat <- as_tibble(FetchData(seurat_obj, vars = "seurat_clusters"), rownames = "cellID")
  annotations_inferCNV <- as_tibble(FetchData(seurat_obj, vars = "infercnv_annotation"), rownames = "cellID")

  # Save annotation files without column names
  write_tsv(annotations_seurat, paste0(output_dir, "annotations_seurat_inferCNV_figure_", sample_name, ".tsv"), col_names = FALSE)
  write_tsv(annotations_inferCNV, paste0(output_dir, "annotations_sample_vs_control_inferCNV_figure_", sample_name, ".tsv"), col_names = FALSE)

  # Print completion message
  print(paste("Gene expression and annotation files saved for:", sample_name))
  
  if(all(colnames(GEX_matrix) %in% annotations_seurat$cellID)){
    print("all cells in matrix are in the annotation data")
  }
  # Return processed data as a list (optional)
  return(GEX_matrix)
}

```

```{r}
run_inferCNV <- function(gene_expression_matrix, annotation_file, gene_order_file, output_dir = "~/results/inferCNV/", ref_groups) {
  
  # Create InferCNV object
  print("create infercnv object")
  inferCNVObj <- CreateInfercnvObject(
    raw_counts_matrix = gene_expression_matrix,
    gene_order_file = gene_order_file,
    annotations_file = annotation_file,
    delim = '\t',
    ref_group_names = ref_groups
  )
  
  # Run InferCNV
  print("run inferCNV")
  inferCNV_res <- infercnv::run(
    inferCNVObj, 
    cutoff = 1, 
    out_dir = output_dir,
    cluster_by_groups = TRUE,
    denoise = TRUE,
    HMM = FALSE,  # Set to TRUE if using Hidden Markov Model for CNV calling
    useRaster = FALSE,
    png_res = 900,
    #output_format = "pdf"
    output_format = "png",
    num_threads = 12
    )
  
  # Print completion message
  print(paste("InferCNV completed and results saved in:", output_dir))
  
  # Return the InferCNV result object
  return(inferCNV_res)
}

```

# subset seurat obj
InferCNV cannot handle all ~ 100000 cells in the dataset. so we will subset the same proportion of cells for each cluster
```{r}
set.seed(1234)
```

```{r}
srt_merge <- readRDS("~/processed_data/tumors_normalized_scaled_dimreduction_annotated_refined_miller_with_malig_states_srt_revision.rds")
```


```{r}
meta <- srt_merge@meta.data
meta$cell <- rownames(meta)

target_n <- 15000
sample_proportion <- target_n/nrow(meta)

sampled_cells <- meta %>% # take metadata
  group_by(seurat_clusters) %>%   # group by samples
  slice_sample(prop = sample_proportion) %>% # take a random proportion of each cluster
  pull(cell)

srt_merge_sub <- subset(srt_merge, cells = sampled_cells)
srt_merge_sub
```

```{r}
# create a new metadata column that labels cells as either sample name or Nonmalignant
srt_merge_sub@meta.data <- srt_merge_sub@meta.data %>% 
  mutate(Malignant = case_when(cell_type_manual == "Malignant" ~ "Malignant",
                               TRUE ~ "Nonmalignant")) 

srt_merge_sub@meta.data$Malignant %>% table()
srt_merge_sub$cell_type_manual %>% table()
# divide nonmalignant into "glia", "lymphoid", "myeloid", and "other
meta <- srt_merge_sub@meta.data

infercnv_annotation <- meta %>%
  mutate(infercnv_annotation = case_when(
    Malignant == "Nonmalignant" & cell_type_manual == "Glial cell" ~ "Glial cell",
    Malignant == "Nonmalignant" & cell_type_manual == "Lymphoid" ~ "Lymphoid",
    Malignant == "Nonmalignant" & cell_type_manual == "Myeloid" ~ "Myeloid",
    Malignant == "Malignant" ~ sample_name,
    TRUE ~ "Other"
    )) %>%
  dplyr::pull(infercnv_annotation)

# check that the vector looks right
infercnv_annotation %>% table()

# add to metadata
srt_merge_sub <- AddMetaData(srt_merge_sub, infercnv_annotation, col.name = "infercnv_annotation")

srt_merge_sub$infercnv_annotation %>% table()
srt_merge_sub$seurat_clusters %>% table()
dim(srt_merge_sub)
```

# create annotation files and GEX matrix for each sample
```{r}
out_dir <- "/your/path/to/resources/annotations_sample_vs_control_tumor_cell_annoation.tsv"

# Process annotations
GEX_matrix <- process_seurat_annotations(
  seurat_obj = srt_merge_sub,
  sample_name = "tumor_cell_annoation", # this is different from the sample name input into run infercnv. This changes the suffix of the annotation files
  output_dir = out_dir)
```

# run InferCNV with fixed malignant cells
cluster by sample
```{r}
# set up resources
annotations_file <- "/your/path/to/resources/annotations_sample_vs_control_tumor_cell_annoation.tsv" # this file is produced by process_seurat_annotations
out_dir_base <- "~/your/path/inferCNV/"
gof <- "~/your/path/resources/hg38_gencode_v27.txt"

sample_name <- "tumor_cell_inferCNV_res"

# important diagnostic! Check that all annotations cells are in the GEX matrix!
annotations <- read_tsv(annotations_file, col_names = FALSE)
all(colnames(GEX_matrix) %in% annotations$X1)
ref_groups <- c("Lymphoid", "Myeloid", "Glial cell", "Other")
annotations$X2 %>% table()

res <- run_inferCNV(gene_expression_matrix = GEX_matrix,
             annotation_file = annotations_file,
             gene_order_file = gof,
             output_dir = paste0(out_dir_base, sample_name),
             ref_groups = ref_groups)
```

