---
title: "Figure3_InferCNV"
output: html_document
date: "2025-04-19"
---

In this script, inferCNV is used to confirm malignant clusters identified by combined UMAP and cellranger cell type calls. seurat objects with dim reductions are imported (script 3) along with a gene order file produced in script 4. Output is inferCNV plots.

Here I am using the gene order file suggested by the infercnv documentation on https://github.com/broadinstitute/inferCNV/wiki/File-Definitions.
Download for this was: https://data.broadinstitute.org/Trinity/CTAT/cnv/
The file is: 	hg38_gencode_v27.txt

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(readr)
library(infercnv)
```


# Define functions
This outputs annotations files for cell ranger cell ids and seurat clusters and returns a gene expression matrix
```{r}
# Define function to process each Seurat object with user-defined clusters
process_seurat_annotations <- function(seurat_obj, sample_name,
                                       output_dir = "~/resources/") {
  
  # Ensure output directory exists
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Extract gene expression matrix
  GEX_matrix <- as.matrix(GetAssayData(seurat_obj, assay = "RNA", slot = "counts"))
  
  # Create annotations for Cell Ranger and Seurat clusters
  annotations_seurat <- as_tibble(FetchData(seurat_obj, vars = "seurat_clusters"), rownames = "cellID")
  annotations_inferCNV <- as_tibble(FetchData(seurat_obj, vars = "infercnv_annotation"), rownames = "cellID")

  # Save annotation files without column names
  write_tsv(annotations_seurat, paste0(output_dir, "annotations_seurat_inferCNV_figure_", sample_name, ".tsv"), col_names = FALSE)
  write_tsv(annotations_inferCNV, paste0(output_dir, "annotations_sample_vs_control_inferCNV_figure_", sample_name, ".tsv"), col_names = FALSE)

  # Print completion message
  print(paste("Gene expression and annotation files saved for:", sample_name))
  
  if(all(colnames(GEX_matrix) %in% annotations_seurat$cellID)){
    print("all cells in matrix are in the annotation data")
  }
  # Return processed data as a list (optional)
  return(GEX_matrix)
}

```



```{r}
run_inferCNV <- function(gene_expression_matrix, annotation_file, gene_order_file, output_dir = "~/results/inferCNV/", ref_groups) {
  
  # Create InferCNV object
  print("create infercnv object")
  inferCNVObj <- CreateInfercnvObject(
    raw_counts_matrix = gene_expression_matrix,
    gene_order_file = gene_order_file,
    annotations_file = annotation_file,
    delim = '\t',
    ref_group_names = ref_groups
  )
  
  # Run InferCNV
  print("run inferCNV")
  inferCNV_res <- infercnv::run(
    inferCNVObj, 
    cutoff = 1, 
    out_dir = output_dir,
    cluster_by_groups = TRUE,
    denoise = TRUE,
    HMM = FALSE,  # Set to TRUE if using Hidden Markov Model for CNV calling
    useRaster = FALSE,
    png_res = 900,
    #output_format = "pdf"
    output_format = "png"
    )
  
  # Print completion message
  print(paste("InferCNV completed and results saved in:", output_dir))
  
  # Return the InferCNV result object
  #return(inferCNV_res)
}

```


```{r}
# inferCNV is largely deterministic but set seed just in case.
set.seed(1234)
```

#import filtered/preprocessed seurat objects
```{r}
#srt_merge <- read_rds("~/processed_data/tumors_normalized_merged_scaled_dimreduction_annotated_srt.rds")
srt_merge <- readRDS("~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/processed_data/tumors_normalized_merged_scaled_dimreduction_annotated_srt.rds")
```

# prep seurat obj
```{r}
# Filter out p16
srt_merge <- subset(srt_merge, orig.ident != "P16_Post_11907")

# for formality, remove redo from orig ident and colnames
srt_merge$orig.ident <- gsub("_redo", "", srt_merge$orig.ident)
colnames(srt_merge) <- gsub("_redo", "", colnames(srt_merge))

# fix sample labels for aesthetics. create as new labels for safety
new_labels <- gsub("_[0-9]{5}$", "", srt_merge$orig.ident)
new_labels <- gsub("_", " ", new_labels)

srt_merge <- AddMetaData(srt_merge, new_labels, col.name = "sample_name")

# Check new names
srt_merge$orig.ident %>% table()
colnames(srt_merge)[srt_merge$orig.ident == "P7_Post_11517"][1:10]
srt_merge$sample_name %>% table()

```

# subset seurat obj
```{r}
# InferCNV cannot handle all ~ 65000 cells in the dataset. so we will subset the same proportion of cells for each cluster
meta <- srt_merge@meta.data
meta$cell <- rownames(meta)

target_n <- 15000
sample_proportion <- target_n/nrow(meta)

sampled_cells <- meta %>% # take metadata
  group_by(seurat_clusters) %>%   # group by clusters
  slice_sample(prop = sample_proportion) %>% # take a random proportion of each cluster
  pull(cell)

srt_merge_sub <- subset(srt_merge, cells = sampled_cells)
```

# Read in malignant calls and annotate seurat object
```{r}
# Read malignant cluster calls
#malig_file <- "~/resources/malignant_cell_calls/tumor_merged.txt"
malig_file <- "~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/resources/malignant_cell_calls/tumor_merged.txt"
malig_clusters <- read_lines(malig_file)
# Get number of clusters
nClusters <- length(table(FetchData(srt_merge_sub, vars = "seurat_clusters")))
  
# Determine non-malignant clusters
nonmalig_clusters <- as.character(setdiff(0:(nClusters - 1), malig_clusters))
```


# create infercnv annotations in metadata to separate by
```{r}
# create a new metadata column that labels cells as either sample name or Nonmalignant
srt_merge_sub@meta.data <- srt_merge_sub@meta.data %>% 
  mutate(malignant = case_when(seurat_clusters %in% malig_clusters ~ "malignant",
                               TRUE ~ "nonmalignant")) 

# divide nonmalignant into "glia", "lymphoid", "myeloid", and "other
srt_merge_sub$cell_type_manual %>% table()
meta <- srt_merge_sub@meta.data

infercnv_annotation <- meta %>%
  mutate(infercnv_annotation = case_when(
    malignant == "nonmalignant" & cell_type_manual == "Glial cell" ~ "Glial cell",
    malignant == "nonmalignant" & cell_type_manual == "Lymphoid" ~ "Lymphoid",
    malignant == "nonmalignant" & cell_type_manual == "Myeloid" ~ "Myeloid",
    malignant == "malignant" ~ orig.ident,
    TRUE ~ "Other"
    )) %>%
  dplyr::pull(infercnv_annotation)

# check that the vector looks right
infercnv_annotation %>% table()

# add to metadata
srt_merge_sub <- AddMetaData(srt_merge_sub, infercnv_annotation, col.name = "infercnv_annotation")

srt_merge_sub$infercnv_annotation %>% table()
dim(srt_merge_sub)
```


# create annotation files and GEX matrix for each sample
```{r}
out_dir <- "~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/resources/"
# Process annotations
GEX_matrix <- process_seurat_annotations(
  seurat_obj = srt_merge_sub,
  sample_name = "tumors_merge", # this is different from the sample name input into run infercnv. This changes the suffix of the annotation files
  output_dir = out_dir)
```


# run InferCNV for each sample

```{r}
# by sample

# set up resurces
#out_dir <- "~/resources"
#annotations_file <- paste0(out_dir,"annotations_sample_vs_control_inferCNV_tumors_merge.tsv")

#out_dir_base <- "~/results/inferCNV/"
#gof <- "~/external_resources/hg38_gencode_v27.txt"


sample_name <- "tumors_merge_by_sample_figure"
annotations_file <- "~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/resources/annotations_sample_vs_control_inferCNV_figure_tumors_merge.tsv"
out_dir_base <- "~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/results/inferCNV/"
gof <- "~/Dropbox (PennNeurosurgery)/SilverbushLab/Projects/E20_tumor_analysis/external_resources/hg38_gencode_v27.txt"

# important diagnostic! Check that all annotations cells are in the GEX matrix!
annotations <- read_tsv(annotations_file, col_names = FALSE)
all(colnames(GEX_matrix) %in% annotations$X1)
ref_groups <- c("Lymphoid", "Myeloid", "Glial cell", "Other")

run_inferCNV(gene_expression_matrix = GEX_matrix,
             annotation_file = annotations_file,
             gene_order_file = gof,
             output_dir = paste0(out_dir_base, sample_name),
             ref_groups = ref_groups)
```



