---
title: "E20_Script6_Manual_Annotation"
output: html_document
date: "2025-04-01"
---

```{r}
# for presto:
#devtools::install_github("immunogenomics/presto")
```


```{r}
library(Seurat)
library(readr)
library(tibble)
library(dplyr)
library(viridis)
library(presto) # used to greatly improve FindAllMarkers efficiency
library(biomaRt)
```

# define funcitons
```{r}
# Define a function to run UMAP
create_UMAP <- function(seurat_obj, sample_name, num_pcs, res = 0.25, umap_dir = "~/results/dimreduction/umap/", processed_data_dir = "~/processed_data/") {

  # Find neighbors and clusters
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs)
  seurat_obj <- FindClusters(seurat_obj, resolution = res)
  
  # Print cluster assignments for verification
  print(paste("Clusters for", sample_name, ":"))
  print(head(Idents(seurat_obj)))
  
  # Run UMAP
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs)

  # Save UMAP plots with different groupings
  p1 <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + ggtitle(sample_name)
  #ggsave(filename = file.path(umap_dir, paste0("UMAP_", sample_name, "_seurat_clusters.jpg")),
         #plot = p1, dpi = 150, height = 10, width = 10)

  #p2 <- DimPlot(seurat_obj, reduction = "umap", group.by = "cell_type_manual_refined", label = TRUE) + ggtitle(sample_name)
  #ggsave(filename = file.path(umap_dir, paste0("UMAP_", sample_name, "_cell_type_manual_refined")),
  #       plot = p2, dpi = 150, height = 10, width = 12)
  print(p1)

  # Print message indicating completion
  print(paste("UMAP and clustering completed for:", sample_name))

  # Return the updated Seurat object
  return(seurat_obj)
}
```

# function to plot markers
```{r}
plot_marker_umaps <- function(seurat_obj, marker_csv, samplename = "sample", path_out) {
  # Load marker gene list
  marker_df <- read_csv(marker_csv)
  
  # Validate columns
  if (!all(c("cell_type", "marker") %in% colnames(marker_df))) {
    stop("CSV must contain 'cell_type' and 'marker' columns")
  }
  
  # Drop NA or blank marker entries
  marker_df <- marker_df[!is.na(marker_df$marker) & marker_df$marker != "", ]
  
  plot_list <- list()

  # First: Add UMAP with Seurat cluster identity
  cluster_plot <- DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5) +
    ggtitle("Clusters")
  plot_list[[1]] <- cluster_plot

  # Then: Add FeaturePlots for each marker
  for (i in seq_len(nrow(marker_df))) {
    marker <- marker_df$marker[i]
    cell_type <- marker_df$cell_type[i]
    
    if (marker %in% rownames(seurat_obj)) {
      p <- FeaturePlot(seurat_obj, features = marker, reduction = "umap", pt.size = 0.25, order = TRUE) +
        ggtitle(paste0(cell_type, ": ", marker))
    } else {
      p <- ggplot() + 
        theme_void() + 
        ggtitle(paste0(cell_type, ": ", marker, " (not found)"))
    }
    plot_list[[i + 1]] <- p  # shift index by 1 to account for cluster plot
  }
  
  # Layout: adjust ncol to control grid width
  ncol <- 6
  nrow <- ceiling(length(plot_list) / ncol)
  final_plot <- wrap_plots(plotlist = plot_list, ncol = ncol)
  
  # Save
  jpg_filename <- paste0(path_out, '/', samplename, ".jpg")
  ggsave(jpg_filename, final_plot, width = 4 * ncol, height = 3 * nrow)
  
  message("Saved plot to: ", jpg_filename)
  return(final_plot)
}
```

```{r}
sample_name <- "tumors_merged"
```

#import filtered/preprocessed seurat objects
```{r}
#srt_merge <- readRDS("~/processed_data/tumors_normalized_merged_scaled_dimreduction_srt_submission.rds"))
```

```{r}
markers_path <- "~/resources/cell_type_markers_for_multi_marker_UMAP.csv"

plot_marker_umaps(srt_merge, markers_path, samplename = "UMAP_annotation_tumors_merged_", path_out = "~/results")
```


# import manual cluster-cell type annotations
```{r}
annotation_path_base <- "~/resources/manual_annotations/"
annotations.df <- read_csv(paste0(annotation_path_base, "annotation_", sample_name, "_submission.csv"))

clusters <- srt_merge$seurat_clusters
celltypes <- annotations.df$cell_type[match(clusters, annotations.df$seurat_cluster)] # use the match function to add manually called cell types for each cluster to each cell.
srt_merge <- AddMetaData(srt_merge, celltypes, col.name = "cell_type_manual")

srt_merge@meta.data$cell_type_manual %>% table()
```


# Annotation dotplot for all cells
```{r }
# output directory
srt_merge@meta.data$cell_type_manual %>% table()
out_dir <- "~/results/annotation/"

# Read in the marker list once
markers_df <- read_csv("~/resources/cell_type_markers_forDotplot_general.csv")
marker_genes <- unique(markers_df$marker)
marker_genes <- factor(marker_genes, levels = marker_genes)

# Define the desired cell type order
desired_order <- c("Malignant", "Neuron", "Glial cell", "Lymphoid", "Myeloid", "Endothelial", "Mixed lymphoid-myeloid")
srt_merge@meta.data$cell_type_manual <- factor(srt_merge@meta.data$cell_type_manual, levels = desired_order)

plot <- DotPlot(srt_merge,  features = marker_genes, group.by= 'cell_type_manual', dot.scale = 6, cluster.idents = FALSE) + RotatedAxis()
#ggsave("~/results/annotation/DotPlot_AllCells_GeneralMarkers.jpg", plot = plot)
plot
```

# Subcluster Lymphoid cells
```{r}
srt_merge_lymphoid <- subset(srt_merge, subset = cell_type_manual == "Lymphoid")
dim(srt_merge_lymphoid)
```

```{r}
set.seed(1234)

sample_name <- "tumors_merged_lymphoid"
# find variable features
srt_merge_lymphoid <- FindVariableFeatures(srt_merge_lymphoid, selection.method = "vst", nfeatures = 2000)
  
# Scale data 
markers_df <- read_csv("~/resources/Tcell_markers_all.csv")
genes_to_scale <- markers_df$marker

srt_merge_lymphoid <- ScaleData(srt_merge_lymphoid, features = c(VariableFeatures(srt_merge_lymphoid), genes_to_scale)) 

srt_merge_lymphoid <- RunPCA(srt_merge_lymphoid, features = VariableFeatures(object = srt_merge_lymphoid))

ElbowPlot(srt_merge_lymphoid)
```

```{r}
srt_merge_lymphoid <- create_UMAP(srt_merge_lymphoid, sample_name = sample_name, 
        num_pcs = 12, res = 0.35)

srt_merge_lymphoid@meta.data %>% head()
srt_merge_lymphoid@meta.data$cell_type_manual %>% table()
```


```{r}
plot_marker_umaps(srt_merge_lymphoid, marker_csv = "~/resources/Tcell_markers_all.csv", samplename = "UMAP_annotation_tumors_merged_lymphoid", path_out = "~/results")
```

# Add manual Lymphoid annotations
```{r}
sample_name <- "tumors_merged_lymphoid"
annotation_path_base <- "~/resources/manual_annotations/"
annotations.df <- read_csv(paste0(annotation_path_base, "annotation_", sample_name, "_submission.csv"))

clusters <- srt_merge_lymphoid$seurat_clusters
celltypes <- annotations.df$cell_type[match(clusters, annotations.df$seurat_cluster)] # use the match function to add manually called cell types for each cluster to each cell.
srt_merge_lymphoid <- AddMetaData(srt_merge_lymphoid, celltypes, col.name = "cell_type_manual_refined")
srt_merge_lymphoid@meta.data$cell_type_manual_refined %>% table()
```

# Lymphoid cell annotation dot plot
```{r}
# read in dotplot markers
marker_genes <- read_csv("~/resources/Tcell_markers_for_dotplot.csv") %>%
  pull(marker)
marker_genes <- factor(marker_genes, levels = marker_genes)

srt_merge_lymphoid@meta.data$cell_type_manual_refined %>% table()

desired_order <- c("CD8 T cell", "CD4 T cell", "Regulatory T cell", "Cycling lymphoid cell", "NK cell")
srt_merge_lymphoid@meta.data$cell_type_manual_refined <- factor(srt_merge_lymphoid@meta.data$cell_type_manual_refined, levels = desired_order)

plot <- DotPlot(srt_merge_lymphoid,  features = marker_genes, group.by= 'cell_type_manual_refined', dot.scale = 6, cluster.idents = FALSE) + RotatedAxis()
#ggsave("~/results/annotation/DotPlot_Lymphoid_GeneralMarkers.jpg", plot = plot)
plot
```


# update the main seurat object with enhanced T cell annotations

```{r}
# change cell types back to character. Otherwise it is impossible to assign values of 2 factors with different levels together. 
srt_merge$cell_type_manual <- as.character(srt_merge$cell_type_manual)
srt_merge_lymphoid$cell_type_manual_refined <- as.character(srt_merge_lymphoid$cell_type_manual_refined)

# create refined annotations for full seurat object
lymphoid_annotations <- srt_merge_lymphoid@meta.data %>%
  dplyr::select(cell_type_manual_refined) %>%
  rownames_to_column(var = "cell_id")

# initialize the refined cell types in main srt obj
srt_merge$cell_type_manual_refined <- srt_merge$cell_type_manual 

# get the cells in both srt_merge and srt_merge_lymphoid
matching_cells <- colnames(srt_merge)[colnames(srt_merge) %in% lymphoid_annotations$cell_id]

# get the positions of the common (lymphoid) cells in the lymphoid annotations (same order as lymphoid srt obj)
match_index <- match(matching_cells, lymphoid_annotations$cell_id)

# For the lymphoid cells in srt_merge, change the value of cell_type_manual_refined to the value of cell_type_manual_refined in the lymphoid annotations (refined)
srt_merge$cell_type_manual_refined[matching_cells] <- lymphoid_annotations$cell_type_manual_refined[match_index]



# and plot over UMAP to be sure:
plot <- DimPlot(srt_merge, group.by = "cell_type_manual_refined")
#ggsave(plot = plot, filename = "~/results/dimreduction/umap/UMAP_tumors_merged_celltypes_manual_withLymphoid.jpg")
plot
```


```{r}
DimPlot(srt_merge, group.by = "cell_type_manual_refined")
```


# Subcluster myeloid cells

```{r}
srt_merge_myeloid <- subset(srt_merge, cell_type_manual %in% c("Myeloid"))
```

Reprocess
```{r}
set.seed(1234)

sample_name <- "tumors_merged_myeloid"

# find variable features
srt_merge_myeloid <- FindVariableFeatures(srt_merge_myeloid, selection.method = "vst", nfeatures = 2000)

# Scale data 
#markers_df <- read_csv("~/resources/cell_type_markers_all.csv")

srt_merge_myeloid <- ScaleData(srt_merge_myeloid, features = VariableFeatures(srt_merge_myeloid))

srt_merge_myeloid <- RunPCA(srt_merge_myeloid, features = VariableFeatures(object = srt_merge_myeloid))

plot <- ElbowPlot(srt_merge)
plot


```

From elbow plot, use 10 PCs
```{r}
srt_merge_myeloid <- create_UMAP(srt_merge_myeloid, sample_name = sample_name, 
        num_pcs = 10, res = 0.35)

srt_merge_myeloid@meta.data %>% head()
```

```{r}
plot_marker_umaps(srt_merge_myeloid, "~/resources/Myeloid_markers_all.csv", samplename = "UMAP_annotation_tumors_merged_myeloid", path_out = "~/results/")
```


 
# Add manual myeloid annotations
```{r}
sample_name <- "tumors_merged_myeloid"
annotation_path_base <- "~/resources/manual_annotations/"
annotations.df <- read_csv(paste0(annotation_path_base, "annotation_", sample_name, "_submission.csv"))

clusters <- srt_merge_myeloid$seurat_clusters
celltypes <- annotations.df$cell_type[match(clusters, annotations.df$seurat_cluster)] # use the match function to add manually called cell types for each cluster to each cell.
srt_merge_myeloid <- AddMetaData(srt_merge_myeloid, celltypes, col.name = "cell_type_manual_refined")
srt_merge_myeloid@meta.data$cell_type_manual_refined %>% table()
```


# Plot over UMAP 
```{r}
p <- DimPlot(srt_merge_myeloid, reduction = "umap", group.by = "cell_type_manual_refined")
#ggsave("~/results/dimreduction/umap/UMAP_tumors_merged_myeloid_cell_type_manual_refined.jpg", plot = p)
p
```

# Myeloid cell annotation dot plot
```{r}
# read in dotplot markers
marker_genes <- read_csv("~/resources/Myeloid_markers_for_dotplot.csv") %>%
  pull(marker)
marker_genes <- factor(marker_genes, levels = marker_genes)

srt_merge_myeloid@meta.data$cell_type_manual_refined %>% table()
unique(srt_merge_myeloid$cell_type_manual_refined)

desired_order <- c("Macrophage", "Monocyte", "pDC", "Plasma cell", "Cycling myeloid cell")

srt_merge_myeloid@meta.data$cell_type_manual_refined <- factor(srt_merge_myeloid@meta.data$cell_type_manual_refined, levels = desired_order)

plot <- DotPlot(srt_merge_myeloid,  features = marker_genes, group.by= 'cell_type_manual_refined', dot.scale = 6, cluster.idents = FALSE) + RotatedAxis()
#ggsave("~/results/annotation/DotPlot_Myeloid_GeneralMarkers.jpg", plot = plot, width = 10, height = 5)
plot
```

# transfer annotations to main seurat object
```{r}
srt_merge_myeloid$cell_type_manual_refined <- as.character(srt_merge_myeloid$cell_type_manual_refined)
srt_merge$cell_type_manual_refined <- as.character(srt_merge$cell_type_manual_refined)

myeloid_annotations <- srt_merge_myeloid@meta.data %>% 
  rownames_to_column(var = "cell_id") %>%
  dplyr::select(cell_id, cell_type_manual_refined)

matching_cells <- colnames(srt_merge)[colnames(srt_merge) %in% myeloid_annotations$cell_id]

match_index <- match(matching_cells, myeloid_annotations$cell_id)

srt_merge$cell_type_manual_refined[matching_cells] <- myeloid_annotations$cell_type_manual_refined[match_index]
srt_merge$cell_type_manual_refined %>% table()
```

```{r}
plot <- DimPlot(srt_merge, group.by = "cell_type_manual_refined")
#ggsave(filename = "~/results/dimreduction/umap/UMAP_tumors_merged_celltypes_manual_with_Lymphoid_Myeloid.jpg")
plot
```

# Add useful metadata
```{r}
sample_name <- gsub("_", " ", srt_merge$orig.ident)
srt_merge <- AddMetaData(srt_merge, sample_name, col.name = "sample_name")

patient <- srt_merge@meta.data$sample_name %>% word()
srt_merge <- AddMetaData(srt_merge, patient, col.name = "patient")

Timepoint <- ifelse(grepl("Pre", srt_merge$sample_name), "Pre", "Post")
srt_merge <- AddMetaData(srt_merge, Timepoint, col.name = "Timepoint")
```


# save final annotated rds objects
```{r}
#saveRDS(srt_merge, file = "~/processed_data/tumors_normalized_merged_scaled_dimreduction_annotated_srt_submission.rds")
```

