---
title: "E17_dimreduction_script2"
output: html_document
date: "2024-12-25"
---

Perform dim reduction on myelin depleted and control GBMs from script 1. Input is processed/filtered seurat objects from script 1. Output is dimensional reduction plots and corresponding seurat objects for cell state scoring and plotting. This is SCRIPT 2 of the pipeline.

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(readr)
```


# Run UMAP
```{r}

# Define a function to run UMAP
create_UMAP <- function(seurat_obj, sample_name, num_pcs, res = 0.25, umap_dir = "~/results/dimreduction/umap/", processed_data_dir = "~/processed_data/") {

  # Find neighbors and clusters
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:num_pcs)
  seurat_obj <- FindClusters(seurat_obj, resolution = res)
  
  # Print cluster assignments for verification
  print(paste("Clusters for", sample_name, ":"))
  print(head(Idents(seurat_obj)))
  
  # Run UMAP
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:num_pcs)

  # Save UMAP plots with different groupings
  p1 <- DimPlot(seurat_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
  ggsave(filename = file.path(umap_dir, paste0("UMAP_", sample_name, "_seurat_clusters.jpg")),
         plot = p1, dpi = 150, height = 10, width = 10)

  p2 <- DimPlot(seurat_obj, reduction = "umap", group.by = "cell_type_coarse", label = TRUE)
  ggsave(filename = file.path(umap_dir, paste0("UMAP_", sample_name, "_cell_type_coarse.jpg")),
         plot = p2, dpi = 150, height = 10, width = 16)

  print(p1)

  # Print message indicating completion
  print(paste("UMAP and clustering completed for:", sample_name))

  # Return the updated Seurat object
  return(seurat_obj)
}
```


#import filtered/preprocessed seurat objects
only do this if necessary bc it takes a long time.
```{r}
#srt_merge <- readRDS("~/processed_data/tumors_normalized_merged_scaled_srt_submission.rds")
```


```{r}
# Set seed so UMAP runs the same every time. Do it before PCA just in case.
set.seed(1234)
```


```{r}
# Define results directory
dimreduction_dir <- "~/results/dimreduction/pca/"
sample <- "tumor_merged_filtered_processed"

# Run PCA
srt_merge <- RunPCA(srt_merge, features = VariableFeatures(object = srt_merge))

# Save ElbowPlot
p4 <- ElbowPlot(srt_merge)
print(p4)
```
From the elbowplot, ideal number of dimensions is:
10


```{r}
# use the function for each object
srt_merge <- create_UMAP(srt_merge, "merged_filtered_processed", res = 0.25, num_pcs =  10)
#saveRDS(srt_merge, "~/processed_data/tumors_normalized_merged_scaled_dimreduction_srt_submission.rds")
```

